---
title: "Adding ESPN and 538 Game Predictions to nflfastR Data"
description: |
  Here, we'll look at how to scrape ESPN's and 538's pregame predictions and merge them into nflfastR data
author:
  - name: Jonathan Goldberg
    url: https://twitter.com/gberg1303
date: 08-29-2020
output:
  distill::distill_article:
    self_contained: false
repository_url: "https://github.com/mrcaseb/open-source-football"
categories: 
  - Scraping
  - Game Predicitions
  - nflfastR
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  dpi = 300,
  tidy = 'styler')
```

In this article, we are going to (1) take a look at how to scrape pregame predictions from 538 and ESPN and (2) how to merge those predictions into nflfastR’s dataset. 

Let’s start by loading up the nflfastR data. To save some time, we’re only going to load the schedules for the last two seasons (2018 and 2019). 

```{r, echo=TRUE, include=TRUE}
library(tidyverse)
NFL_Games <- nflfastR::fast_scraper_schedules(2018:2019) %>%
    dplyr::mutate(
      gameday = lubridate::as_date(gameday)
    )
```

## 538: 

538’s data is super easy to grab. They have a repo of NFL games, each team’s ELO ratings before the game, and their predictions for the outcome of the game.  

```{r, echo=TRUE, include=TRUE}
FiveThirtyEight_Predictions <- read_csv("https://projects.fivethirtyeight.com/nfl-api/nfl_elo.csv")
```

The last thing we need to do is to change some of their labels for NFL teams. Washington and Oakland (Las Vegas) are the two problem children.  

```{r, echo=TRUE, include=TRUE}
FiveThirtyEight_Predictions <- FiveThirtyEight_Predictions %>%
  dplyr::mutate(
    team1 = gsub("WSH", "WAS", team1),
    team2 = gsub("WSH", "WAS", team2),
    team1 = gsub("LAR", "LA", team1),
    team2 = gsub("LAR", "LA", team2)) %>%
  dplyr::select(date, season, team1, team2, elo_prob1) %>%
  dplyr::rename(fivethirtyeight_home_wp = elo_prob1)
```

Now we can merge!

```{r, echo=TRUE, include=TRUE}
NFL_Games <- NFL_Games %>%
  left_join(
    FiveThirtyEight_Predictions,
    by = c("home_team" = "team1", "away_team" = "team2", "gameday" = "date", "season")
  )
```

## ESPN:

ESPN is going to be pretty tricky, but they also include moneyline odds from some sportsbooks. It may be worth the extra effort to convert them to probabilities. 

Unfortunately, the data on ESPN’s pregame predictions cannot be simply scraped from their website nor are they in a repo. To complicate this even more, ESPN has a unique identification system that does not follow a reproducible pattern. To get their game IDs, you have to scrape them from the schedule page for each week of the NFL season. 

I’ve built a function that will help with exactly that. 

```{r, echo=TRUE, include=TRUE}
get_game_ids <- function(season, season_type = c("preseason", "regular", "postseason")){
  current_year <- as.double(substr(Sys.Date(), 1, 4))
  espn_game_ids <- data.frame()

  if (!season_type %in% c("preseason", "regular", "postseason", "all")) {
    stop("Please choose season_type of 'regular',  'playoffs', 'postseason', or 'all'")
  }

  if (!dplyr::between(as.numeric(season), 2002, current_year)) {
    stop(paste("Please choose season between 2002 and", current_year))
  }

  if(lubridate::month(Sys.Date()) < 12 & lubridate::month(Sys.Date()) > 2 & season_type == "postseason" & current_year == season | season_type == "postseason" & lubridate::month(Sys.Date()) <= 2 & current_year == season){
    stop(paste("Unfortunately, the NFL Playoff Games have not been determined yet"))
  }

  message(
    dplyr::if_else(
      season_type == "regular",
      glue::glue("Scraping from {season} {season_type} season!"),
      glue::glue("Scraping from {season} {season_type}!")
    )
  )

  season_type <- ifelse(season_type == "preseason", "1", season_type)
  season_type <- ifelse(season_type == "regular", "2", season_type)
  season_type <- ifelse(season_type == "postseason", "3", season_type)

  weeks <- ifelse(season_type == "2", 17, 5)

   espn_game_ids <- purrr::map_df(1:weeks, function(week){
      url <- glue::glue("https://www.espn.com/nfl/schedule/_/week/{week}/year/{season}/seasontype/{season_type}")

      webpage <- xml2::read_html(url)

      links <- webpage %>%
        rvest::html_nodes("a") %>%
        rvest::html_attr("href")

      espn_gameid <- links %>%
        as.tibble() %>%
        dplyr::filter(str_detect(value, "gameId") == TRUE) %>%
        dplyr::pull(value) %>%
        stringr::str_remove(., "/nfl/game/_/gameId/")

      bye_teams <- webpage %>%
        rvest::html_nodes(".odd.byeweek" ) %>%
        rvest::html_nodes("abbr") %>%
        rvest::html_text()

      home_team <- webpage %>%
        rvest::html_nodes(".home-wrapper") %>%
        rvest::html_nodes("abbr") %>%
        rvest::html_text()

      away_team <- webpage %>%
        rvest::html_nodes("abbr") %>%
        rvest::html_text()
      away_team <- away_team[!away_team %in% home_team]
      away_team <- away_team[!away_team %in% bye_teams]

      placeholder <- data.frame(
        home_team,
        away_team,
        espn_gameid
      ) %>%
        dplyr::mutate(
          season_type = season_type,
          season = season,
          week = ifelse(season_type == 3, 17 + week, week)
        )

      espn_game_ids <- dplyr::bind_rows(espn_game_ids, placeholder)
      return(espn_game_ids)

    }
      )

  ### Fix Several Names for Compatibility with nflfastR Data game_ids
  espn_game_ids <- espn_game_ids %>%
    dplyr::mutate(
      home_team = gsub("WSH", "WAS", home_team),
      away_team = gsub("WSH", "WAS", away_team),
      home_team = gsub("LAR", "LA", home_team),
      away_team = gsub("LAR", "LA", away_team)
    ) %>%
  # Add nflfastR game_ids
    dplyr::mutate(
      week = ifelse(week == 22, week-1, week),
      alt_gameid = paste0(season, "_", ifelse(week >= 10, paste0(week), paste0(0,week)), "_", away_team, "_", home_team)
    )

  return(espn_game_ids)
}

```

This is only step one, but pulling the pregame predictions from ESPN’s API is rather easy now that we have the IDs for each game. We simply plug them into the link to their json file, do some cleaning along the way, and extract the prediction. 

```{r, echo=TRUE, include=TRUE}
# Get Game IDs
ESPN_Games <- purrr::map_df(2018:2019, function(x){
  get_game_ids(x, season_type = "regular")
})

head(ESPN_Games)
```

```{r, echo=TRUE, include=TRUE}
# Pull Pregame Predictions
ESPN_Game_Predictions <- purrr::map_df(ESPN_Games$espn_gameid, function(espn_game_id){
  
  pregame_predictions <- data.frame(espn_gameid = espn_game_id)

  # Pull the JSon
  game_json <- httr::GET(url = glue::glue("http://site.api.espn.com/apis/site/v2/sports/football/nfl/summary?event={espn_game_id}")) %>%
    httr::content(as = "text", encoding = "UTF-8") %>%
    jsonlite::fromJSON(flatten = TRUE)


  # Pull the game data from the ID dataframe
  if("gameProjection" %in% names(game_json[["predictor"]][["homeTeam"]]) == TRUE){
  pregame_predictions <- pregame_predictions %>%
    dplyr::mutate(
      espn_home_wp = as.numeric(game_json[["predictor"]][["homeTeam"]][["gameProjection"]])/100
    )
    message(
      paste("Pulling predictions for", pregame_predictions$alt_gameid)
    )
  }


  # Grab and convert the Moneylines from Oddsmakers
  if("pickcenter" %in% names(game_json) == TRUE &
     "provider.name" %in% names(game_json[["pickcenter"]]) == TRUE &
     "homeTeamOdds.moneyLine" %in% names(game_json[["pickcenter"]]) == TRUE
     ){
  vegas_odds <- data.frame(
    providers = game_json[["pickcenter"]][["provider.name"]],
    odds = ifelse(game_json[["pickcenter"]][["homeTeamOdds.moneyLine"]] > 0, 100/(game_json[["pickcenter"]][["homeTeamOdds.moneyLine"]] + 100), game_json[["pickcenter"]][["homeTeamOdds.moneyLine"]]/(game_json[["pickcenter"]][["homeTeamOdds.moneyLine"]]-100))
    ) %>%
    tidyr::pivot_wider(names_from = providers, values_from = odds)

  pregame_predictions <- cbind(
    pregame_predictions, vegas_odds
  )

  }

  return(pregame_predictions)
})

# Merge ESPN Data together
ESPN_Games <- ESPN_Games %>% 
  left_join(
    ESPN_Game_Predictions, 
    by = c("espn_gameid" = "espn_gameid")
  )

# Merge back to main data 
NFL_Games <- NFL_Games %>%
  left_join(
    ESPN_Games %>% select(alt_gameid, espn_home_wp, Caesars, numberfire, teamrankings, consensus),
    by = c("game_id" = "alt_gameid")
  )
```

Now that it’s all together, let’s take a look at the accuracy of ESPN’s, 538’s, Numberfire's predictions over the last two seasons. 

```{r, echo=TRUE, include=TRUE}
### Do some data wrangling first
NFL_Games <- NFL_Games %>%
  mutate(home_win = ifelse(home_score > away_score, 1, 0),
         correct_espn = ifelse(ifelse(espn_home_wp > .5, 1, 0) == home_win, 1, 0),
         correct_numberfire = ifelse(ifelse(numberfire > .5, 1, 0) == home_win, 1, 0),
         correct_fivethirtyeight = ifelse(ifelse(fivethirtyeight_home_wp > .5, 1, 0) == home_win, 1, 0))
Accuracy_Dataset <- NFL_Games  %>%
  # Filter out Playoff Games
  filter(game_type == "REG") %>%
  # Pivot Longer to allow group_by and summarize
  pivot_longer(
    cols = starts_with("correct_"),
    names_to = "predictor",
    names_prefix = "correct_",
    values_to = "correct"
  ) %>%
  # Group and Summarize
  group_by(predictor) %>%
  summarise(
    games = sum(!is.na(correct)),
    games_correct = sum(correct, na.rm = TRUE),
    percent_correct = round(mean(correct, na.rm = TRUE),3)
  )

### Merge Over Brier Scores
Accuracy_Dataset <- Accuracy_Dataset %>%
  left_join(data.frame(
    predictor = c("espn", "fivethirtyeight", "numberfire"),
    brier_score = c(
      DescTools::BrierScore(NFL_Games %>% filter(!is.na(espn_home_wp)) %>% pull(home_win), NFL_Games %>% filter(!is.na(espn_home_wp)) %>% pull(espn_home_wp)),
      DescTools::BrierScore(NFL_Games$home_win, NFL_Games$fivethirtyeight_home_wp),
      DescTools::BrierScore(NFL_Games %>% filter(!is.na(numberfire)) %>% pull(home_win), NFL_Games %>% filter(!is.na(espn_home_wp)) %>% pull(numberfire))
      )),
    by = "predictor"
    ) %>%
  mutate(brier_score = round(brier_score, 3))
```

```{r, layout="l-page", fig.height=5.5, preview=TRUE}
### Create a fancy reactable table
library(reactable)
Accuracy_Dataset %>%
  arrange(-percent_correct) %>%
  reactable(
    compact = TRUE,
    borderless = FALSE,
    striped = FALSE,
    fullWidth = TRUE,
    defaultColDef = colDef(
      align = "center",
      minWidth = 100),
    theme = reactableTheme(
      headerStyle = list(
        "&:hover[aria-sort]" = list(background = "hsl(0, 0%, 96%)"),
        "&[aria-sort='ascending'], &[aria-sort='descending']" = list(background = "hsl(0, 0%, 96%)"),
        borderColor = "#555")
    ),
    columns = list(
      predictor = colDef(name = "Predictor",
                         align = "left",
                          minWidth = 110),
      games = colDef(name = "Predictions"),
      games_correct = colDef(name = "Correct Predictions"),
      percent_correct = colDef(name = "% Correct",
                               class = "border-left",
                               style = function(value) {
                                 normalized <- (value - min(Accuracy_Dataset$percent_correct)) / (max(Accuracy_Dataset$percent_correct) - min(Accuracy_Dataset$percent_correct))
                                 color <- rgb(colorRamp(c("#0089BA","#00B0A7"))(normalized), maxColorValue = 255)
                                 list(background = color)}
                               ),
      brier_score = colDef(name = "Brier Score", 
                           style = function(value) {
                             normalized <- (value - min(Accuracy_Dataset$brier_score)) / (max(Accuracy_Dataset$brier_score) - min(Accuracy_Dataset$brier_score))
                             color <- rgb(colorRamp(c("#0089BA", 	"#00B0A7"))(normalized), maxColorValue = 255)
                             list(background = color)
                             }
                           )
    )
  )
```

It's unsurprising that Numberfire is significantly more accurate than the ESPN or 538 predictions. As an oddsmaker, they can see where bets are placed and update their lines accordingly. Essentially, Numberfire benefits from more information.

I'm looking forward to the future investigations into predicting NFL games that may come from this data!
