---
title: "Exploring Rolling Averages of EPA"
description: |
  Use nflfastR to develop rolling averages of EPA to best predict a game's outcome
author:
  - name: Jack Lichtenstein
    url: https://twitter.com/jacklich10
date: "12/29/20"
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_depth: 3
repository_url: "https://github.com/mrcaseb/open-source-football"
categories:
  - nflfastR
  - EPA
  - Rolling EPA
  - Prediction
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  message = FALSE, 
  warning = FALSE, 
  dpi = 300, 
  tidy = 'styler'
  )
```

This post is inspired by a brief twitter [thread](https://twitter.com/greerreNFL/status/1343830333783822338) between [Lee Sharpe](https://twitter.com/LeeSharpeNFL) and [Robby Greer](https://twitter.com/greerreNFL) as well as Jonathan Goldberg's previous [post](https://www.opensourcefootball.com/posts/2020-08-20-adjusting-epa-for-strenght-of-opponent/) on Open Source Football that adjusts EPA/play for opponent using 10 game rolling windows. The goal of this article is to alter EPA/play by adjusting for opponent as well as to determine the best rolling average window to maximize the predictive power of future game outcomes.

```{r libraries, echo=FALSE, include=FALSE}
# Load libraries
library(tidyverse)
library(nflfastR)
# Initialize team colors
NFL_pri <- c('ARI'='#97233f',
             'ATL'='#a71930',
             'BAL'='#241773',
             'BUF'='#00338d',
             'CAR'='#0085ca',
             'CHI'='#0b162a',
             'CIN'='#000000',
             'CLE'='#fb4f14',
             'DAL'='#002244',
             'DEN'='#002244',
             'DET'='#005a8b',
             'GB'='#203731',
             'HOU'='#03202f',
             'IND'='#002c5f',
             'JAX'='#000000',
             'KC'='#e31837',
             'LAC'='#002244',
             'LA'='#003693',
             'MIA'='#008e97',
             'MIN'='#4f2683',
             'NE'='#002244',
             'NO'='#9f8958',
             'NYG'='#0b2265',
             'NYJ'='#125740',
             'OAK'='#a5acaf',
             'LV'='#a5acaf',
             'PHI'='#004953',
             'PIT'='#000000',
             'SF'='#aa0000',
             'SEA'='#002244',
             'TB'='#d50a0a',
             'TEN'='#002244',
             'WAS'='#773141')

NFL_sec <- c('pos'='#FFFFFF',
             'neg'='#000000',
             'ARI'='#000000',
             'ATL'='#000000',
             'BAL'='#000000',
             'BUF'='#c60c30',
             'CAR'='#000000',
             'CHI'='#c83803',
             'CIN'='#fb4f14',
             'CLE'='#22150c',
             'DAL'='#b0b7bc',
             'DEN'='#fb4f14',
             'DET'='#b0b7bc',
             'GB'='#ffb612',
             'HOU'='#a71930',
             'IND'='#a5acaf',
             'JAX'='#006778',
             'KC'='#ffb612',
             'LAC'='#0073cf',
             'LA'='#ffd000',
             'MIA'='#f58220',
             'MIN'='#ffc62f',
             'NE'='#c60c30',
             'NO'='#000000',
             'NYG'='#a71930',
             'NYJ'='#000000',
             'OAK'='#000000',
             'LV'='#000000',
             'PHI'='#a5acaf',
             'PIT'='#ffb612',
             'SF'='#b3995d',
             'SEA'='#69be28',
             'TB'='#34302b',
             'TEN'='#4b92db',
             'WAS'='#ffb612')
```

# Loading in the data

Let's start by first reading in play-by-play data from 1999-2020 from @nflfastR.

```{r load data, echo=TRUE, include=TRUE}
#tictoc::tic()
nfl_pbp <- purrr::map_df(1999:2020, function(x) {
  readr::read_csv(
    glue::glue("https://raw.githubusercontent.com/guga31bb/nflfastR-data/master/data/play_by_play_{x}.csv.gz")
  ) %>% 
    dplyr::select(-starts_with("blocked_"))
})
tictoc::toc()
```

# Methodology

Just as Jonathan did in his post, we will find every team's weekly EPA/play on offense and defense. Additionally, instead of simply finding total offense and defense EPA/play, we will break it down into passing and rushing EPA/play for offense and defense.

```{r epa calc, echo=TRUE, include=TRUE}
### Get game EPA data
# Offense EPA
epa_data <- nfl_pbp %>%
  dplyr::filter(!is.na(epa), !is.na(ep), !is.na(posteam),
                play_type == "pass" | play_type == "run" | penalty == 1, qb_kneel != 1) %>%
  dplyr::group_by(game_id, season, week, posteam, home_team) %>%
  dplyr::summarise(off_dropback_pct = mean(qb_dropback == 1),
                   off_epa = mean(epa),
                   off_pass_epa = mean(epa[qb_dropback == 1]),
                   off_rush_epa = mean(epa[qb_dropback == 0]),
                   off_epa_n = sum(qb_dropback == 1 | qb_dropback == 0),
                   off_pass_epa_n = sum(qb_dropback == 1),
                   off_rush_epa_n = sum(qb_dropback == 0),
                   .groups = "drop") %>%
  # Defense EPA
  dplyr::left_join(nfl_pbp %>%
                     filter(!is.na(epa), !is.na(ep), !is.na(posteam), 
                            play_type == "pass" | play_type == "run" | penalty == 1, qb_kneel != 1) %>%
                     dplyr::group_by(game_id, season, week, defteam, away_team) %>%
                     dplyr::summarise(def_epa = mean(epa),
                                      def_dropback_pct = mean(qb_dropback == 1),
                                      def_pass_epa = mean(epa[qb_dropback == 1]),
                                      def_rush_epa = mean(epa[qb_dropback == 0]),
                                      def_epa_n = sum(qb_dropback == 1 | qb_dropback == 0),
                                      def_pass_epa_n = sum(qb_dropback == 1),
                                      def_rush_epa_n = sum(qb_dropback == 0),
                                      .groups = "drop"),
                   by = c("game_id", "posteam" = "defteam", "season", "week")) %>%
  dplyr::mutate(opponent = ifelse(posteam == home_team, away_team, home_team)) %>%
  dplyr::select(game_id, season, week, home_team, away_team, posteam, opponent, 
                off_dropback_pct, off_epa, off_pass_epa, off_rush_epa,
                off_epa_n, off_pass_epa_n, off_rush_epa_n,
                def_epa_n, def_pass_epa_n, def_rush_epa_n,
                def_dropback_pct, def_epa, def_pass_epa, def_rush_epa) %>% 
  # Not sure why, but there is one instance where the posteam = ""
  dplyr::filter(posteam != "")
```

```{r remove pbp, include=FALSE, echo=FALSE}
# Remove pbp from environment
rm(nfl_pbp)
```

We are going to build our dataset in a very similar manner to what Jonathan did in his post, with a few changes outlined below:

* Convert each EPA statistic into a lagging moving average. The lag ensures that we compare a team’s performance against their opponent’s performance *up to that point in the season*.

* Instead of weighting each game equally during the window, we will weight EPA by the number of plays in each game of the window.

* Instead of simply converting each statistic into a moving average of the last ten games, we will convert each statistic into a moving average using a dynamic window that ranges from ten games to twenty games (for teams that play in the Super Bowl). In other words, we will use a ten game window to predict the winner of the 11th game, but for say the 15th game, we will use a 14 game window. 

* Note that for new seasons, this ten game window serves as a prior for each team in a similar manner to Football Outsiders' weighted DVOA. For example, to predict week 3, we will use a rolling average that includes 8 games from the previous year along with weeks 1 and 2. 

* We will ignore the 1999 season and only analyze the predictive power of EPA starting with the 2000 season, allowing us to use the 1999 season as a prior.

* Finally, the `pracma` package offers several types of [rolling averages](https://rdrr.io/rforge/pracma/man/movavg.html), including simple, weighted, running and exponential rolling averages. We will play around with these types of moving averages to see what maximizes predictive power.

If you want to look at all the code for building this dataset, view the source file! Here is the function to compute moving averages based on varying window sizes:

```{r weight moving avg func, echo=TRUE, include=TRUE}
# Function to get moving average of a dynamic window from 10 - 20 games
  wt_mov_avg_local <- function(var, weight, window, type, moving = T) {
    if (length(weight) == 1 & weight[1] == 1) {
      weight <- rep(1, length(var))
    }
    if (moving) {
      dplyr::case_when(
        window == 10 ~ pracma::movavg(var*weight, n = 10, type = type)/
          pracma::movavg(weight, n = 10, type = type),
        window == 11 ~ pracma::movavg(var*weight, n = 11, type = type)/
          pracma::movavg(weight, n = 11, type = type),
        window == 12 ~ pracma::movavg(var*weight, n = 12, type = type)/
          pracma::movavg(weight, n = 12, type = type),
        window == 13 ~ pracma::movavg(var*weight, n = 13, type = type)/
          pracma::movavg(weight, n = 13, type = type),
        window == 14 ~ pracma::movavg(var*weight, n = 14, type = type)/
          pracma::movavg(weight, n = 14, type = type),
        window == 15 ~ pracma::movavg(var*weight, n = 15, type = type)/
          pracma::movavg(weight, n = 15, type = type),
        window == 16 ~ pracma::movavg(var*weight, n = 16, type = type)/
          pracma::movavg(weight, n = 16, type = type),
        window == 17 ~ pracma::movavg(var*weight, n = 17, type = type)/
          pracma::movavg(weight, n = 17, type = type),
        window == 18 ~ pracma::movavg(var*weight, n = 18, type = type)/
          pracma::movavg(weight, n = 18, type = type),
        window == 19 ~ pracma::movavg(var*weight, n = 19, type = type)/
          pracma::movavg(weight, n = 19, type = type),
        window == 20 ~ pracma::movavg(var*weight, n = 20, type = type)/
          pracma::movavg(weight, n = 20, type = type)
      )
    } else {
      pracma::movavg(var*weight, n = 10, type = type)/
        pracma::movavg(weight, n = 10, type = type)
    }
  }
```

```{r functions, echo=FALSE, include=FALSE}
# Function to create the dataset
create_rolling_data <- function(epa_data, move = T, team_n = 10, pt_diff_type = "r", epa_off = "e", epa_pass_o = "e", epa_rush_o = "r", epa_def = "s", epa_pass_d = "s", epa_rush_d = "r", epa_drop = "r") {
  
  # Join back opponent off/def EPA
  epa_data <- epa_data %>%
    dplyr::group_by(season, posteam) %>% 
    dplyr::mutate(game_number = row_number()) %>% 
    dplyr::left_join(epa_data %>%
                       dplyr::group_by(season, posteam) %>% 
                       dplyr::mutate(opp_game_number = row_number()) %>% 
                       dplyr::select(-opponent) %>%
                       dplyr::rename(opp_off_epa = off_epa,
                                     opp_off_pass_epa = off_pass_epa,
                                     opp_off_rush_epa = off_rush_epa,
                                     opp_def_epa = def_epa,
                                     opp_def_pass_epa = def_pass_epa,
                                     opp_def_rush_epa = def_rush_epa,
                                     opp_off_epa_n =  off_epa_n, 
                                     opp_off_pass_epa_n = off_pass_epa_n, 
                                     opp_off_rush_epa_n = off_rush_epa_n,
                                     opp_def_epa_n =  def_epa_n, 
                                     opp_def_pass_epa_n = def_pass_epa_n, 
                                     opp_def_rush_epa_n = def_rush_epa_n,
                                     opp_off_dropback_pct = off_dropback_pct,
                                     opp_def_dropback_pct = def_dropback_pct) %>%
                       dplyr::group_by(posteam) %>%
                       dplyr::arrange(season, week) %>%
                       dplyr::mutate(
                         window = ifelse(opp_game_number <= 10, team_n, opp_game_number),
                         # Opponent off EPA
                         opp_off_epa = dplyr::lag(wt_mov_avg_local(var = opp_off_epa, weight = opp_off_epa_n, window = window, type = epa_off, moving = move)),
                         opp_off_pass_epa = dplyr::lag(wt_mov_avg_local(var = opp_off_pass_epa, weight = opp_off_pass_epa_n, window = window, type = epa_pass_o, moving = move)),
                         opp_off_rush_epa = dplyr::lag(wt_mov_avg_local(var = opp_off_rush_epa, weight = opp_off_rush_epa_n, window = window, type = epa_rush_o, moving = move)),
                         # Opponent def EPA
                         opp_def_epa = dplyr::lag(wt_mov_avg_local(var = opp_def_epa, weight = opp_def_epa_n, window = window, type = epa_def, moving = move)),
                         opp_def_pass_epa = dplyr::lag(wt_mov_avg_local(var = opp_def_pass_epa, weight = opp_def_pass_epa_n, window = window, type = epa_pass_d, moving = move)),
                         opp_def_rush_epa = dplyr::lag(wt_mov_avg_local(var = opp_def_rush_epa, weight = opp_def_rush_epa_n, window = window, type = epa_rush_d, moving = move)),
                         # Opponent defense dropbacks
                         opp_def_dropback_pct = dplyr::lag(wt_mov_avg_local(var = opp_def_dropback_pct, weight = 1, window = window, type = epa_drop, moving = move)),
                         # Opponent offense dropbacks
                         opp_off_dropback_pct = dplyr::lag(wt_mov_avg_local(var = opp_off_dropback_pct, weight = 1, window = window, type = epa_drop, moving = move))),
                     by = c("game_id", "season", "week", "home_team", "away_team", "opponent" = "posteam")) %>% 
    # Fix errors that occur for "1999_01_BAL_STL", "2000_06_BUF_MIA", "2000_03_SD_KC" games (NAs)
    dplyr::mutate(dplyr::across(c(opp_off_dropback_pct:opp_off_rush_epa, opp_def_dropback_pct:opp_def_rush_epa),
                                ~ ifelse(is.na(.) & week != 1, 0, .)))
  
  # Join in league mean of EPA by season, week to prepare for opponent adjustments
  epa_data <- epa_data %>%
    dplyr::left_join(epa_data %>%
                       dplyr::filter(posteam == home_team) %>%
                       dplyr::group_by(season, week) %>%
                       dplyr::summarise(league_mean_total = sum(off_epa*off_epa_n + def_epa*def_epa_n)/sum(off_epa_n + def_epa_n),
                                        league_mean_dropback_pct = (mean(off_dropback_pct) + mean(def_dropback_pct))/2,
                                        league_mean_pass = sum(off_pass_epa*off_pass_epa_n + def_pass_epa*def_pass_epa_n)/sum(off_pass_epa_n + def_pass_epa_n),
                                        league_mean_rush = sum(off_rush_epa*off_rush_epa_n + def_rush_epa*def_rush_epa_n)/sum(off_rush_epa_n + def_rush_epa_n),
                                        .groups = "drop") %>%
                       dplyr::ungroup() %>%
                       dplyr::group_by(season) %>%
                       dplyr::mutate(league_mean_total = dplyr::lag(cummean(league_mean_total)),
                                     league_mean_dropback_pct = dplyr::lag(cummean(league_mean_dropback_pct)),
                                     league_mean_pass = dplyr::lag(cummean(league_mean_pass)),
                                     league_mean_rush = dplyr::lag(cummean(league_mean_rush))) %>% 
                       dplyr::ungroup(),
                     by = c("season", "week")) %>% 
    dplyr::ungroup()
  
  # Adjust EPA for opponent
  epa_data <- epa_data %>%
    dplyr::mutate(
      # Total off/def adjustments
      off_adjustment_factor = ifelse(!is.na(league_mean_total), league_mean_total-opp_def_epa, 0),
      def_adjustment_factor = ifelse(!is.na(league_mean_total), league_mean_total-opp_off_epa, 0),
      adjusted_off_epa = off_epa + off_adjustment_factor,
      adjusted_def_epa = def_epa + def_adjustment_factor,
      # Dropback pct off/def adjustments
      off_dropback_adjustment_factor = ifelse(!is.na(league_mean_dropback_pct), league_mean_dropback_pct-opp_def_dropback_pct, 0),
      def_dropback_adjustment_factor = ifelse(!is.na(league_mean_dropback_pct), league_mean_dropback_pct-opp_off_dropback_pct, 0),
      adjusted_off_dropback_pct = off_dropback_pct + off_dropback_adjustment_factor,
      adjusted_def_dropback_pct = def_dropback_pct + def_dropback_adjustment_factor,
      # Pass off/def adjustments
      off_pass_adjustment_factor = ifelse(!is.na(league_mean_pass), league_mean_pass-opp_def_pass_epa, 0),
      def_pass_adjustment_factor = ifelse(!is.na(league_mean_pass), league_mean_pass-opp_off_pass_epa, 0),
      adjusted_off_pass_epa = off_pass_epa + off_pass_adjustment_factor,
      adjusted_def_pass_epa = def_pass_epa + def_pass_adjustment_factor,
      # Rush off/def adjustments
      off_rush_adjustment_factor = ifelse(!is.na(league_mean_rush), league_mean_rush-opp_def_rush_epa, 0),
      def_rush_adjustment_factor = ifelse(!is.na(league_mean_rush), league_mean_rush-opp_off_rush_epa, 0),
      adjusted_off_rush_epa = off_rush_epa + off_rush_adjustment_factor,
      adjusted_def_rush_epa = def_rush_epa + def_rush_adjustment_factor)
  
  # Group and calculate rolling average of EPA metrics
  epa_data <- epa_data %>%
    dplyr::group_by(posteam) %>%
    dplyr::arrange(season, week) %>%
    dplyr::mutate(
      window = ifelse(game_number <= 10, team_n, game_number),
      ### Current metrics
      # Total off/def epa
      off_epa_curr = wt_mov_avg_local(var = off_epa, weight = off_epa_n, window = window, type = epa_off, moving = move),
      def_epa_curr = wt_mov_avg_local(var = def_epa, weight = def_epa_n, window = window, type = epa_def, moving = move),
      adjusted_off_epa_curr = wt_mov_avg_local(var = adjusted_off_epa, weight = off_epa_n, window = window, type = epa_off, moving = move),
      adjusted_def_epa_curr = wt_mov_avg_local(var = adjusted_def_epa, weight = def_epa_n, window = window, type = epa_def, moving = move),
      # Pass off/def epa
      off_pass_epa_curr = wt_mov_avg_local(var = off_pass_epa, weight = off_pass_epa_n, window = window, type = epa_pass_o, moving = move),
      def_pass_epa_curr = wt_mov_avg_local(var = def_pass_epa, weight = def_pass_epa_n, window = window, type = epa_pass_d, moving = move),
      adjusted_off_pass_epa_curr = wt_mov_avg_local(var = adjusted_off_pass_epa, weight = off_pass_epa_n, window = window, type = epa_pass_o, moving = move),
      adjusted_def_pass_epa_curr = wt_mov_avg_local(var = adjusted_def_pass_epa, weight = def_pass_epa_n, window = window, type = epa_pass_d, moving = move),
      # Rush off/def epa
      off_rush_epa_curr = wt_mov_avg_local(var = off_rush_epa, weight = off_rush_epa_n, window = window, type = epa_rush_o, moving = move),
      def_rush_epa_curr = wt_mov_avg_local(var = def_rush_epa, weight = def_rush_epa_n, window = window, type = epa_rush_d, moving = move),
      adjusted_off_rush_epa_curr = wt_mov_avg_local(var = adjusted_off_rush_epa, weight = off_rush_epa_n, window = window, type = epa_rush_o, moving = move),
      adjusted_def_rush_epa_curr = wt_mov_avg_local(var = adjusted_def_rush_epa, weight = def_rush_epa_n, window = window, type = epa_rush_d, moving = move),
      # Dropback pct
      off_dropback_pct_curr = wt_mov_avg_local(var = off_dropback_pct, weight = 1, window = window, type = epa_drop, moving = move),
      def_dropback_pct_curr = wt_mov_avg_local(var = def_dropback_pct, weight = 1, window = window, type = epa_drop, moving = move),
      adjusted_off_dropback_pct_curr = wt_mov_avg_local(var = adjusted_off_dropback_pct, weight = 1, window = window, type = epa_drop, moving = move),
      adjusted_def_dropback_pct_curr = wt_mov_avg_local(var = adjusted_def_dropback_pct, weight = 1, window = window, type = epa_drop, moving = move),
      ### Lagged metrics
      dplyr::across(c(off_epa_curr:adjusted_def_dropback_pct_curr),
                    ~ dplyr::lag(.),
                    .names = "{.col}_{.fn}")) %>% 
    dplyr::select(-c(off_epa, def_epa, adjusted_off_epa, adjusted_def_epa,
                     off_pass_epa, def_pass_epa,
                     adjusted_off_pass_epa, adjusted_def_pass_epa,
                     off_rush_epa, def_rush_epa,
                     adjusted_off_rush_epa, adjusted_def_rush_epa,
                     off_dropback_pct, def_dropback_pct,
                     adjusted_off_dropback_pct, adjusted_def_dropback_pct)) %>% 
    dplyr::rename_with(.cols = ends_with("_1"), ~ str_remove(., "_curr_1")) %>% 
    dplyr::ungroup() %>%
    dplyr::select(game_id, season, week, posteam,
                  off_epa, def_epa, adjusted_off_epa, adjusted_def_epa,
                  off_dropback_pct, def_dropback_pct, adjusted_off_dropback_pct, adjusted_def_dropback_pct,
                  off_pass_epa, def_pass_epa, adjusted_off_pass_epa, adjusted_def_pass_epa,
                  off_rush_epa, def_rush_epa, adjusted_off_rush_epa, adjusted_def_rush_epa,
                  ends_with("_curr"))
  
  ### Get schedule and game outcomes from Lee Sharpe
  weekly_outcomes <- suppressWarnings(readr::read_csv("https://raw.githubusercontent.com/leesharpe/nfldata/master/data/games.csv", col_types = cols())) %>% 
    dplyr::filter(!is.na(result)) %>% 
    dplyr::mutate_at(dplyr::vars(home_team, away_team),
                     ~ stringr::str_replace_all(., c("JAC" = "JAX",
                                                     "STL" = "LA",
                                                     "SL" = "LA",
                                                     "ARZ" = "ARI",
                                                     "BLT" = "BAL",
                                                     "CLV" = "CLE",
                                                     "HST" = "HOU",
                                                     "SD" = "LAC",
                                                     "OAK" = "LV")))
  
  # Double games (one row per team rather than one row per game)
  weekly_outcomes <- weekly_outcomes %>% 
    dplyr::transmute(season, week, game_date = gameday, game_id,
                     home_team, away_team, home_score, away_score,
                     team = away_team,
                     opponent = home_team,
                     points_for = away_score,
                     points_against = home_score,
                     point_differential = -result) %>% 
    dplyr::bind_rows(., weekly_outcomes %>% 
                       dplyr::transmute(season, week, game_date = gameday, game_id,
                                        home_team, away_team, home_score, away_score,
                                        team = home_team,
                                        opponent = away_team,
                                        points_for = home_score,
                                        points_against = away_score,
                                        point_differential = result)) %>% 
    dplyr::mutate(win = ifelse(point_differential > 0, 1, 0),
                  winner = ifelse(point_differential > 0, team, opponent),
                  loser = ifelse(point_differential < 0, team, opponent)) %>% 
    dplyr::arrange(season, week) %>% 
    dplyr::group_by(season, team) %>% 
    dplyr::mutate(game_number = dplyr::row_number()) %>% 
    dplyr::ungroup() %>% 
    dplyr::relocate(c(game_number, team, opponent, winner, loser), .after = game_id)
  
  # Join back opponent game outcome metrics
  weekly_outcomes <- weekly_outcomes %>%
    dplyr::left_join(weekly_outcomes %>%
                       dplyr::select(game_id, season, week, game_number, team, opponent, points_for, points_against) %>%
                       dplyr::rename(opp_points_for = points_for, 
                                     opp_points_against = points_against,
                                     opp_game_number = game_number) %>%
                       dplyr::group_by(team) %>%
                       dplyr::arrange(season, week) %>%
                       dplyr::mutate(window = ifelse(opp_game_number <= 10, team_n, opp_game_number),
                                     opp_points_for = dplyr::lag(wt_mov_avg_local(var = opp_points_for, weight = 1, window = window, type = pt_diff_type, moving = move)),
                                     opp_points_against = dplyr::lag(wt_mov_avg_local(var = opp_points_against, weight = 1, window = window, type = pt_diff_type, moving = move))) %>%
                       dplyr::select(-season, -week),
                     by = c("game_id", "team" = "opponent", "opponent" = "team"))
  
  # Join league average point scoring, differentials for opponent adjustments
  weekly_outcomes <- weekly_outcomes %>% 
    dplyr::left_join(weekly_outcomes %>%
                       dplyr::filter(team == home_team) %>%
                       dplyr::group_by(season, week) %>%
                       dplyr::summarise(league_mean_pts = mean(points_for),
                                        .groups = "drop") %>%
                       dplyr::ungroup() %>%
                       dplyr::group_by(season) %>%
                       dplyr::mutate(league_mean_pts = dplyr::lag(cummean(league_mean_pts))),
                     by = c("season", "week")) %>% 
    # Adjust points for
    dplyr::mutate(off_adjustment_factor = ifelse(!is.na(league_mean_pts) & !is.na(opp_points_against), league_mean_pts-opp_points_against, 0),
                  def_adjustment_factor = ifelse(!is.na(league_mean_pts) & !is.na(opp_points_for), league_mean_pts-opp_points_for, 0),
                  adjusted_points_for = points_for + off_adjustment_factor,
                  adjusted_points_against = points_against + def_adjustment_factor,
                  adjusted_point_differential = adjusted_points_for - adjusted_points_against)
  
  # Group and calculate rolling average of point differential metrics
  weekly_outcomes <- weekly_outcomes %>%
    dplyr::group_by(team) %>%
    dplyr::arrange(season, week) %>%
    dplyr::mutate(
      window = ifelse(game_number <= 10, team_n, game_number),
      ### Current metrics
      adjusted_points_for_curr = wt_mov_avg_local(var = adjusted_points_for, weight = 1, window = window, type = pt_diff_type, moving = move),
      adjusted_points_against_curr = wt_mov_avg_local(var = adjusted_points_against, weight = 1, window = window, type = pt_diff_type, moving = move),
      point_differential_curr = wt_mov_avg_local(var = point_differential, weight = 1, window = window, type = pt_diff_type, moving = move),
      adjusted_point_differential_curr = wt_mov_avg_local(var = adjusted_point_differential, weight = 1, window = window, type = pt_diff_type, moving = move),     
      ### Lagged metrics
      across(c(adjusted_points_for_curr:adjusted_point_differential_curr),
             ~ dplyr::lag(.),
             .names = "{.col}_{.fn}")) %>%
    dplyr::select(-c(point_differential, adjusted_point_differential,
                     adjusted_points_for, adjusted_points_against)) %>% 
    dplyr::rename_with(.cols = ends_with("_1"), ~ str_remove(., "_curr_1")) %>% 
    dplyr::ungroup() %>% 
    dplyr::select(-c(league_mean_pts, off_adjustment_factor, def_adjustment_factor))
  
  ### Create Model Dataset
  model_dataset <- weekly_outcomes %>%
    # Add opponent box score statistics
    dplyr::left_join(weekly_outcomes %>%
                       dplyr::select(game_id, team, adjusted_points_for, adjusted_points_against,
                                     point_differential, adjusted_point_differential) %>%
                       dplyr::rename(opp_point_differential = point_differential, 
                                     opp_adjusted_points_for = adjusted_points_for,
                                     opp_adjusted_points_against = adjusted_points_against,
                                     opp_adjusted_point_differential = adjusted_point_differential),
                     by = c("game_id", "opponent" = "team")) %>%
    # Add EPA statistics
    dplyr::left_join(epa_data, by = c("game_id", "season", "week", "home_team" = "posteam")) %>%
    dplyr::left_join(epa_data %>%
                       dplyr::rename(
                         # Total off/def EPA
                         opp_off_epa = off_epa,
                         opp_def_epa = def_epa,
                         opp_adjusted_off_epa = adjusted_off_epa,
                         opp_adjusted_def_epa = adjusted_def_epa,
                         # Dropback off/def pct
                         opp_off_dropback_pct = off_dropback_pct,
                         opp_def_dropback_pct = def_dropback_pct,
                         opp_adjusted_off_dropback_pct = adjusted_off_dropback_pct,
                         opp_adjusted_def_dropback_pct = adjusted_def_dropback_pct,
                         # Pass off/def EPA
                         opp_off_pass_epa = off_pass_epa,
                         opp_def_pass_epa = def_pass_epa,
                         opp_adjusted_off_pass_epa = adjusted_off_pass_epa,
                         opp_adjusted_def_pass_epa = adjusted_def_pass_epa,
                         # Rush off/def EPA
                         opp_off_rush_epa = off_rush_epa,
                         opp_def_rush_epa = def_rush_epa,
                         opp_adjusted_off_rush_epa = adjusted_off_rush_epa,
                         opp_adjusted_def_rush_epa = adjusted_def_rush_epa),
                     by = c("game_id", "season", "week", "away_team" = "posteam")) %>%
    dplyr::filter(home_team == team) %>%
    # Add home margin
    dplyr::mutate(home_margin = home_score - away_score) %>%
    # Filter NAs
    dplyr::filter(!is.na(off_epa)) %>%
    # Add numeric ID
    dplyr::mutate(numeric_id = row_number()) %>% 
    dplyr::rename(gameday = game_date)

  return(model_dataset)
}

# Function to compute predictiveness
double_for_accuracy_check <- function(dataset, type) {
  # Double games (one row per team rather than one row per game)
  g1 <- dataset %>% 
    dplyr::transmute(gameday, game_id, season, week, 
                     team = away_team,
                     opponent = home_team,
                     win = ifelse(win == 1, 0, 1),
                     margin = -home_margin,
                     opp_adjusted_point_differential,
                     opp_adjusted_off_epa,
                     opp_adjusted_def_epa,
                     opp_adjusted_off_pass_epa,
                     opp_adjusted_def_pass_epa,
                     opp_adjusted_off_rush_epa,
                     opp_adjusted_def_rush_epa,
                     opp_adjusted_off_dropback_pct,
                     opp_adjusted_def_dropback_pct,
                     opp_point_differential,
                     opp_off_epa,
                     opp_def_epa,
                     opp_off_pass_epa,
                     opp_def_pass_epa,
                     opp_off_rush_epa,
                     opp_def_rush_epa,
                     opp_off_dropback_pct,
                     opp_def_dropback_pct) %>% 
    dplyr::rename_with(.cols = starts_with("opp_"), ~ stringr::str_remove(., "opp_")) %>% 
    dplyr::mutate(location = "Away")
  
  g2 <- dataset %>% 
    dplyr::transmute(gameday, game_id, season, week, 
                     team = home_team,
                     opponent = away_team,
                     win,
                     margin = home_margin,
                     adjusted_point_differential,
                     adjusted_off_epa,
                     adjusted_def_epa,
                     adjusted_off_pass_epa,
                     adjusted_def_pass_epa,
                     adjusted_off_rush_epa,
                     adjusted_def_rush_epa,
                     adjusted_off_dropback_pct,
                     adjusted_def_dropback_pct,
                     point_differential,
                     off_epa,
                     def_epa,
                     off_pass_epa,
                     def_pass_epa,
                     off_rush_epa,
                     def_rush_epa,
                     off_dropback_pct,
                     def_dropback_pct) %>% 
    dplyr::mutate(location = "Home")
  
  doubled <- dplyr::bind_rows(g1, g2) %>% 
    dplyr::arrange(game_id, gameday, season, week) %>% 
    dplyr::mutate(off_pass_rush_epa = off_dropback_pct*off_pass_epa + (1-off_dropback_pct)*off_rush_epa,
                  def_pass_rush_epa = def_dropback_pct*def_pass_epa + (1-adjusted_def_dropback_pct)*adjusted_def_rush_epa,
                  adjusted_off_pass_rush_epa = adjusted_off_dropback_pct*adjusted_off_pass_epa + (1-adjusted_off_dropback_pct)*adjusted_off_rush_epa,
                  adjusted_def_pass_rush_epa = adjusted_def_dropback_pct*adjusted_def_pass_epa + (1-adjusted_def_dropback_pct)*adjusted_def_rush_epa)
  
  results <- doubled %>% 
    dplyr::filter(season > 1999) %>% 
    tidyr::pivot_longer(cols = c(-c(gameday:margin, location)), 
                        names_to = "metric", 
                        values_to = "value") %>% 
    tidyr::nest(data = c(-metric)) %>% 
    dplyr::mutate(regression = map(data, ~ glm(win ~ value, data = ., family = "binomial")),
                  r_squared = map(regression, fmsb::NagelkerkeR2)) %>% 
    tidyr::hoist(r_squared, r.squared = "R2") %>% 
    dplyr::arrange(desc(r.squared)) %>% 
    dplyr::select(metric, r.squared) %>% 
    dplyr::mutate(moving_avg = type)
}

# Function to plot results
plot_results <- function(results) {
  results %>% 
    dplyr::mutate(name = stringr::str_to_title(stringr::str_replace_all(metric, "_", " ")),
                  name = stringr::str_remove(name, "usted"),
                  type = dplyr::case_when(
                    stringr::str_detect(metric, "pass_rush") ~ "Complex",
                    stringr::str_detect(metric, "pass") ~ "Pass",
                    stringr::str_detect(metric, "rush") ~ "Rush",
                    stringr::str_detect(metric, "off_epa|def_epa") ~ "Total",
                    stringr::str_detect(metric, "point_diff") ~ "Point Diff",
                    stringr::str_detect(metric, "dropback") ~ "Dropback")) %>%
    dplyr::mutate(type = forcats::fct_relevel(type, "Point Diff", "Total", "Complex", "Pass", "Rush", "Dropback"),
                  name = forcats::fct_reorder(name, r.squared),
                  moving_avg = forcats::fct_reorder(moving_avg, r.squared)) %>% 
    dplyr::filter(type != "Dropback") %>% 
    ggplot(aes(r.squared, name, fill = moving_avg)) +
    geom_col(position = position_dodge(0.9)) +
    geom_text(aes(label = round(r.squared, 4)),
              hjust = 1, size = 2.5, color = "white", fontface = "bold",
              position = position_dodge(0.9)) +
    facet_wrap( ~ type, scales = "free") +
    scale_x_continuous(expand = c(0, 0)) +
    theme_bw() +
    theme(plot.title = element_text(face = "bold", size = 28/.pt, hjust = 0),
          plot.subtitle = element_text(face = "italic", size = 24/.pt),
          strip.background = element_rect(color = "black", fill = "#C0C0C0", size = 3.5, linetype = "blank"),
          strip.text.x = element_text(face = "bold"),
          panel.grid.minor.x = element_blank(),
          panel.grid.minor.y = element_blank(),
          panel.grid.major.y = element_blank(),
          panel.border = element_blank(),
          axis.ticks = element_blank(),
          axis.text = element_text(size = 24/.pt),
          axis.title = element_text(face = "bold", size = 26/.pt),
          plot.caption = element_text(face = "italic", size = 20/.pt)) +
    labs(x = expression(bold("Nagelkerke pseudo"~R^2)),
         y = NULL,
         fill = NULL,
         title = "Predictive power of various metrics for the outcome of an NFL game",
         subtitle = "Logistic regression to predict win/loss",
         caption = "Chart: @jacklich10 | Data: @nflfastR | Games from 2000-2020 NFL seasons")
}
```

# Analyzing predictive power

First, let's create a dataset where we use *simple* moving averages along with a *static* window of just ten games as a baseline (this is the idea in Jonathan's post) and compare it with a dataset where we use a *simple* moving averages along with a *dynamic* window that ranges from ten to twenty games.

```{r static dynamic, include=FALSE, echo=FALSE}
# Simple static window
model_dataset <- create_rolling_data(epa_data, move = F, pt_diff_type = "s", 
                                     epa_off = "s", epa_pass_o = "s", epa_rush_o = "s", 
                                     epa_def = "s", epa_pass_d = "s", epa_rush_d = "s", 
                                     epa_drop = "s")

results_simple <- double_for_accuracy_check(dataset = model_dataset, type = "Simple")
# Dynamic window
model_dataset <- create_rolling_data(epa_data, move = T, pt_diff_type = "s", 
                                     epa_off = "s", epa_pass_o = "s", epa_rush_o = "s", 
                                     epa_def = "s", epa_pass_d = "s", epa_rush_d = "s", 
                                     epa_drop = "s")

results_dynamic <- double_for_accuracy_check(dataset = model_dataset, type = "Dynamic")

results <- dplyr::bind_rows(results_simple, results_dynamic)
```

```{r plot init, layout="l-page", fig.height=5}
# Plot
plot_results(results) +
  scale_fill_brewer(palette = "Dark2", 
                      guide = guide_legend(reverse = T))
```

Awesome! It looks like using the dynamic window (which always adds the additional information from all games played during a season) increases the predictive power of EPA/play, rather than using a static window that looks only at a team's previous ten games.

Note that the "Complex" category combines pass and rush EPA/play together as follows:

$$\text{off_pass_rush_epa} = \text{off_dropback_pct}*\text{off_pass_epa} + (1-\text{off_dropback_pct})*\text{off_rush_epa}$$

In other words, it combines a team's pass and rush EPA/play into a total EPA/play by weighting them by the percentage of time a team drops back to pass. The difference between this and total EPA/play is that the complex version adjusts pass and rush EPA separately according to the strength of opponent pass and rush EPA, whereas total EPA adjusts based on opponent holistically.

What's interesting is that adjusting defensive EPA/play for opponent offenses faced actually *decreases* its predictive power, albeit a very small amount. This finding might provide a little support to the "defenses are simply a product of the offenses they face", as penalizing or rewarding defenses based on the quality of the offense they face might simply be adding noise. Of course, in general, defensive strength as measured through EPA is much less predictive than offensive strength. 

Now that we know that using a dynamic window for the rolling averages works better than a simple static ten game window, let's play around with different types of moving averages from the `pracma` package. We will try *weighted*, *running* and *exponential* averages compared to the *simple* moving averages from above.

```{r pracma explore, include=FALSE, echo=FALSE}
# Weighted
model_dataset <- create_rolling_data(epa_data, move = T, pt_diff_type = "w", 
                                     epa_off = "w", epa_pass_o = "w", epa_rush_o = "w", 
                                     epa_def = "w", epa_pass_d = "w", epa_rush_d = "w", 
                                     epa_drop = "w")

results_weighted <- double_for_accuracy_check(dataset = model_dataset, type = "Weighted")
# Runnning
model_dataset <- create_rolling_data(epa_data, move = T, pt_diff_type = "r", 
                                     epa_off = "r", epa_pass_o = "r", epa_rush_o = "r", 
                                     epa_def = "r", epa_pass_d = "r", epa_rush_d = "r", 
                                     epa_drop = "r")

results_running <- double_for_accuracy_check(dataset = model_dataset, type = "Running")
# Exponential
model_dataset <- create_rolling_data(epa_data, move = T, pt_diff_type = "e", 
                                     epa_off = "e", epa_pass_o = "e", epa_rush_o = "e", 
                                     epa_def = "e", epa_pass_d = "e", epa_rush_d = "e", 
                                     epa_drop = "e")

results_exp <- double_for_accuracy_check(dataset = model_dataset, type = "Exponential")

results <- dplyr::bind_rows(results_dynamic %>% 
                              dplyr::mutate(moving_avg = "Simple"), results_weighted, results_running, results_exp)
```

```{r plot pracma, layout="l-page", fig.height=5}
plot_results(results) +
  scale_fill_brewer(palette = "Dark2", 
                      guide = guide_legend(reverse = T))
```

There is a lot to take in here, so let's break it down by category of predictor:

* Point Differential: It seems clear that a running moving average of point differential increases its predictive power the most.

* Total Offense/Defense EPA: In terms of offensive EPA, an exponential moving average works best (this is primarily driven by passing offense). In terms of defensive EPA, simple or running moving averages work best and it is difficult to discern between the two.

* Offense Pass EPA: An exponential moving average best predicts offensive pass EPA.

* Defense Pass EPA: Simple and running moving averages do about equally.

* Offense/Defense Rush EPA: A running moving average outclasses the other options.

In general, weighting recent performance more heavily increases a metric's predictive strength. Let's see if we can't combine the best types of rolling average for each predictor to generate the best set of predictors for the outcome of a game. We will use a running average for point differential, rushing offense/defense, as well as for passing and total defense and an exponential moving average for offensive pass EPA and total offensive EPA.

```{r mixed, include=FALSE, echo=FALSE}
# Mixed moving averages
model_dataset <- create_rolling_data(epa_data, move = T, pt_diff_type = "r", 
                                     epa_off = "e", epa_pass_o = "e", epa_rush_o = "r", 
                                     epa_def = "r", epa_pass_d = "r", epa_rush_d = "r", 
                                     epa_drop = "r")

results_mixed <- double_for_accuracy_check(dataset = model_dataset, type = "Mixed")

results <- dplyr::bind_rows(results_dynamic %>% 
                              dplyr::mutate(moving_avg = "Simple"), 
                            results_weighted, results_running, results_exp, results_mixed)
```

```{r plot mixed, layout="l-page", fig.height=5}
plot_results(results) +
  scale_fill_brewer(palette = "Dark2", 
                      guide = guide_legend(reverse = T))
```

So mixing the types of moving averages yields some benefits and some drawbacks. We are able to achieve the maximum level of predictive power for point differential and rushing offense/defense, however passing offense/defense suffers. It appears that when adjusting passing offense for opponent, it is better to use a lagging exponential moving average of opponent defense, yet this lagging exponential moving average of defense significantly decreases the predictive power of defensive EPA. Obviously, more work can be done to play around with different combinations of moving averages, although some improvements could easily be attributed to random noise. For now, let's use the mixed dataset to explore!

# Exploration

If you follow [Ben Baldwin](https://twitter.com/benbbaldwin) on twitter or have visited his awesome website [https://rbsdm.com/](rbsdm.com), you may be familiar with his weekly team tiers scatter plot. Let's look at the differences between those team tiers (which use unadjusted, cumulative mean EPA) and team tiers using our framework:

```{r bind adj unadj, include=FALSE, echo=FALSE}
# Join unadjusted EPA/play together with moving average, adjusted version
current_epa <- dplyr::bind_rows(model_dataset %>% 
                                  dplyr::filter(season == 2020) %>% 
                                  dplyr::filter(week == max(week)) %>% 
                                  dplyr::transmute(team,
                                                   adj_off_epa = adjusted_off_epa_curr.x,
                                                   adj_def_epa = adjusted_def_epa_curr.x),
                                model_dataset %>% 
                                  dplyr::filter(season == 2020) %>% 
                                  dplyr::filter(week == max(week)) %>% 
                                  dplyr::transmute(team = opponent,
                                                   adj_off_epa = adjusted_off_epa_curr.y,
                                                   adj_def_epa = adjusted_def_epa_curr.y)) %>% 
  dplyr::left_join(epa_data %>% 
                     dplyr::filter(season == 2020) %>% 
                     dplyr::group_by(team = posteam) %>% 
                     dplyr::summarise(off_epa = sum(off_epa*off_epa_n)/sum(off_epa_n),
                                      def_epa = sum(def_epa*def_epa_n)/sum(def_epa_n)) %>% 
                     dplyr::ungroup(),
                   by = "team") %>% 
  dplyr::left_join(nflfastR::teams_colors_logos %>% 
                     dplyr::select(team = team_abbr, team_logo_espn),
                   by = "team")
```

```{r team tiers, layout="l-page", fig.height=5, preview=TRUE}
# Plot unadjusted team tiers with adjusted team tiers
current_epa %>% 
  ggplot() +
  geom_segment(aes(x = off_epa, xend = adj_off_epa,
                   y = def_epa, yend = adj_def_epa, color = team),
               lineend = "butt", linejoin = "round", size = 1.75, alpha = 0.8,
               arrow = arrow(length = unit(0.03, "npc"), type = "closed")) +
  geom_point(aes(off_epa, def_epa, color = team),
             size = 2.5) +
  ggimage::geom_image(aes(adj_off_epa, adj_def_epa, image = team_logo_espn),
                      asp = 1.618, by = "height", size = 0.075) +
  # Averages as dashed lines
  geom_vline(aes(xintercept = mean(adj_off_epa)), lty = 2, color = "black") +
  geom_hline(aes(yintercept = mean(adj_def_epa)), lty = 2, color = "black") +
  # Slopes for 'tiers'
  geom_abline(slope = -1.5, intercept = c(-0.5, -0.4, -0.3, -0.2, -0.1, 0, 0.1, 0.2, 0.3, 0.4, 0.5), alpha = .2) +
  coord_equal() +
  scale_y_reverse() +
  scale_color_manual(values = NFL_pri) +
  theme_bw() +
  theme(aspect.ratio = 9/16,
        legend.position = "none",
        plot.title = element_text(face = "bold", size = 28/.pt, hjust = 0),
        plot.subtitle = element_text(face = "italic", size = 24/.pt),
        strip.background = element_rect(color = "black", fill = "#C0C0C0", size = 3.5, linetype = "blank"),
        strip.text.x = element_text(face = "bold"),
        panel.border = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(size = 24/.pt),
        axis.title = element_text(face = "bold", size = 26/.pt),
        plot.caption = element_text(face = "italic", size = 20/.pt)) +
  labs(title = "2020 Predictive NFL Team Tiers after week 16",
       subtitle = "Arrows point from unadjusted, cumulative mean EPA to adjusted and predictive EPA",
       x = "Offensive EPA",
       y = "Defensive EPA",
       caption = "Chart: @jacklich10 | Data: @nflfastR")
```

Wow, look at the Bills and Ravens! Everybody thinks the Chiefs are unbeatable, but their path to the Super Bowl seems much more treacherous... We also see that the Rams and Steelers offenses and defenses have regressed since the beginning of the year.

As a Jets fan, I can't help but look at the Jets ruining their quest for Trevor Lawrence as the Jags ensure the losing continues! `r emo::ji("cry")`

```{r efficiency func, include=FALSE, echo=FALSE}
# Function to display a team's efficiency over time
efficiency_over_time <- function(dataset, current_season, current_week, team_name) {
  # Double games (one row per team rather than one row per game)
  g1 <- dataset %>% 
    dplyr::transmute(gameday, game_id, season, week, 
                     team = away_team,
                     opponent = home_team,
                     adjusted_off_epa = opp_adjusted_off_epa,
                     adjusted_def_epa = opp_adjusted_def_epa) %>% 
    dplyr::mutate(location = "Away")
  
  g2 <- dataset %>% 
    dplyr::transmute(gameday, game_id, season, week, 
                     team = home_team,
                     opponent = away_team,
                     adjusted_off_epa,
                     adjusted_def_epa) %>% 
    dplyr::mutate(location = "Home")
  
  # Bind together and fill in offense/defense efficiency for bye weeks
  doubled <- dplyr::bind_rows(g1, g2) %>% 
    dplyr::arrange(game_id, gameday, season, week) %>% 
    tidyr::complete(season, week, team) %>% 
    dplyr::mutate(season = ifelse(week == 1, season - 1, season),
                  week = ifelse(week == 1, 17, week - 1)) %>% 
    dplyr::group_by(team) %>% 
    tidyr::fill(adjusted_off_epa, adjusted_def_epa, .direction = "updown") %>% 
    dplyr::group_by(season, week) %>% 
    dplyr::mutate(off_rank = rank(-adjusted_off_epa, ties.method = "random"),
                  def_rank = rank(adjusted_def_epa, ties.method = "random")) %>% 
    dplyr::ungroup() %>% 
    dplyr::left_join(nflfastR::teams_colors_logos %>%
                       select(team_abbr, team_logo_espn), 
                     by = c("team" = "team_abbr")) %>% 
    #dplyr::filter(week <= 17, is.na(gameday)) %>%
    dplyr::filter(season %in% current_season, week <= current_week) %>% 
    tidyr::pivot_longer(cols = ends_with("_rank"), names_to = "side", values_to = "rank") %>% 
    dplyr::mutate(side = ifelse(side == "off_rank", "Offensive Efficiency", "Defensive Efficiency"),
                  side = fct_relevel(side, "Offensive Efficiency", "Defensive Efficiency"))
  
  if (length(current_season) > 1) {
    season <- paste0(current_season[1], "-", current_season[length(current_season)])
  } else {
    season <- current_season
  }
  
  plot <- doubled %>% 
    ggplot(aes(week, rank, color = team)) +
    geom_line(aes(size = ifelse(team == team_name, "A", "B"), alpha = ifelse(team == team_name, 1, 0.2)), 
              show.legend = F) +
    ggimage::geom_image(aes(max(week), rank, image = ifelse(week == max(week[team == team_name]) & team == team_name, team_logo_espn, NA)),
                        asp = 1.618, by = "height", size = 0.15, inherit.aes = F) +
    facet_wrap(~ side, nrow = 1) +
    scale_x_continuous(breaks = seq(0, 17, by = 1)) +
    scale_y_reverse(breaks = seq(1, 32, by = 3)) +
    scale_color_manual(values = NFL_pri) +
    scale_size_manual(values = c(1.5, 0.75)) +
    theme_bw() +
    theme(aspect.ratio = 9/16,
          plot.title = element_text(face = "bold", size = 28/.pt, hjust = 0),
          plot.subtitle = element_text(face = "italic", size = 24/.pt),
          strip.background = element_rect(color = "black", fill = "#C0C0C0", size = 3.5, linetype = "blank"),
          strip.text.x = element_text(face = "bold"),
          panel.grid.minor.x = element_blank(),
          panel.grid.minor.y = element_blank(),
          panel.grid.major.x = element_blank(),
          panel.border = element_blank(),
          axis.ticks = element_blank(),
          axis.text = element_text(size = 24/.pt),
          axis.title = element_text(face = "bold", size = 26/.pt),
          axis.title.y = element_text(angle = 0, vjust = 0.5),
          plot.caption = element_text(face = "italic", size = 20/.pt)) +
    labs(title = paste0(season, " ", team_name, " Efficiency"),
         subtitle = "EPA/Play adjusted for opponent",
         x = "Week",
         y = "League\nRank",
         caption =  "Chart: @jacklich10 | Data: @nflfastR")
  
  if (length(current_season) > 1) {
    return(plot + facet_wrap(season ~ side))
  } else {
    return(plot + facet_wrap(~ side, nrow = 1))
  }
}
```

We can also look at a team's offensive and defensive efficiency over the course of the season using our methodology. Here's the Chiefs and some AFC competitors:

```{r afc, layout="l-page", fig.height=5}
chiefs <- efficiency_over_time(model_dataset, current_season = 2020, current_week = 16, team_name = "KC")
bills <- efficiency_over_time(model_dataset, current_season = 2020, current_week = 16, team_name = "BUF")
ravens <- efficiency_over_time(model_dataset, current_season = 2020, current_week = 16, team_name = "BAL")

cowplot::plot_grid(chiefs, bills, ravens, nrow = 3)
```

And here's some NFC competitors:

```{r nfc, layout="l-page", fig.height=5}
packers <- efficiency_over_time(model_dataset, current_season = 2020, current_week = 16, team_name = "GB")
saints <- efficiency_over_time(model_dataset, current_season = 2020, current_week = 16, team_name = "NO")
bucs <- efficiency_over_time(model_dataset, current_season = 2020, current_week = 16, team_name = "TB")

cowplot::plot_grid(packers, saints, bucs, nrow = 3)
```

# Final Thoughts

The main objective of this post was to improve the *predictiveness* of EPA by exploring rolling averages and dynamically moving windows. Certainly, rolling 250 play averages of EPA, for example, can provide important *descriptive* information. However, as with most predictive measures, this work provides evidence that using the widest window of available data yields the best results. Nonetheless, there are definitely ways to improve this methodology. More work can be done in altering which types of moving averages are used for each facet of play. Additionally, note that the post does not explore weighting EPA for given types of plays. For example, weighting EPA by win probability, increasing the value of EPA on early downs or down-weighting turnovers will probably yield significant increases in its predictive power. It is also somewhat surprising that adjusting pass and rush EPA individually for opponent does not seem to produce much of an advantage, although perhaps a more in-depth technique might yield predictive benefits. Hopefully this analysis is another small iteration in improving EPA as a predictive metric!

Thanks to Sebastian Carl and Ben Baldwin for creating this forum and providing awesome open source football analysis throughout the year, as well as to Jonathan Goldberg, Lee Sharpe and Robby Greer for their inspiration.

```{r gh-source, results='asis', echo=FALSE}
'%>%' <- magrittr::`%>%`
fld <- fs::path_wd() %>% fs::path_split() %>% purrr::pluck(1) %>% tibble::as_tibble() %>% dplyr::slice_tail(n = 1)
fn <- fs::path_wd() %>% fs::dir_ls() %>% fs::path_filter("*.Rmd") %>% fs::path_rel()
glue::glue('<a href="https://github.com/mrcaseb/open-source-football/blob/master/_posts/{fld}/{fn}"
               style="font-family:Consolas;color:blue;background-color:#f8f8f8;align:right;font-size:75%;"
              >View source code on GitHub
           </a>'
           )
```


