---
title: "Gunslingers and Game Managers"
description: |
  Quantifying Quarterbacks' performance along the "game manager" and "gunslinger" dimensions. 
author:
  - name: David Anderson
    url: https://twitter.com/DAnderso13
date: "`r Sys.Date()`"
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_depth: 3
repository_url: "https://github.com/mrcaseb/open-source-football"
categories:
  - Quarterback Analysis
  - nflfastR
  - Stupid Internet Arguments
---

<!-- ####################################################################### -->
<!-- Please keep the following chunk as is at the top of your document. 
     It will set some global chunk options.  -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  message = FALSE, 
  warning = FALSE, 
  dpi = 300, 
  tidy = 'styler'
  )
```

<!-- ####################################################################### -->


Gunslingers vs. Game Managers is the stupidest football argument out there. So, naturally, I love it. 
This post is going to quantify what gunslingers are, what game managers are, and measure these for every QB since 2005. 

Gunslingers are higher risk, higher reward type of quarterbacks - typically they throw for a ton of yards, have a lot of touchdown passes, scramble and run around a lot, throw lots of deep balls, but also take negative plays, and throw interceptions. Brett Favre is the patron saint of gunslingers (unfortunately his prime gunslinging days predate the nflfastR pbp data), but players like Jameis Winston and Rex "Screw It, I'm Going Deep" Grossman are also known as gunslingers, as well.  

Game Managers are lower-risk (and potentially also lower-reward) players. A game manager doesn't make mistakes - takes very few sacks, doesn't throw interceptions and completes a percent of their passes. 

First, we load our data:

```{r}
library(pacman)
p_load(tidyverse, janitor)
seasons <- 2005:2020

pbp <- purrr::map_df(seasons, function(x) {
  readRDS(
    url(
      glue::glue("https://raw.githubusercontent.com/guga31bb/nflfastR-data/master/data/play_by_play_{x}.rds")
    )
  )
})


```


Now I just want the plays that are the QB, so all dropbacks.  I am excluding designed QB runs as those exist outside the gunslinger vs game manager paradigm. I'm also going to strip out the season by taking the first 4 digits of the game id. 

```{r}
pbp %>% filter(qb_dropback == 1, !is.na(epa)) -> passes

passes <- passes %>% mutate(year = substr(game_id,1,4))
```


Now we want to summarize each QB's performance each year: 
What is a Game Manager? A game manager doesn't throw interceptions, doesn't take sacks, doesn't take negative plays (which we'll measure by success rate), and complete a lot of passes. 

```{r}
Stats <- passes %>% filter(home_wp > 0.1, home_wp < 0.9, qb_dropback==1) %>% 
           mutate(QB = ifelse(is.na(passer_player_name), rusher_player_name, passer_player_name)) %>% 
           group_by(QB, year) %>% 
           summarize(numDropbacks = n(),
                    passAttempts = sum(pass_attempt),
                    #Game Manager Calculations:
                    completions = sum(complete_pass),
                    completionPercent = completions / passAttempts,
                    ints = sum(interception),
                    sacks = sum(sack)/numDropbacks,
                    successRate = mean(success),
                    
                    #Gunslinger Calculations:
                    meanAirYards = mean(air_yards, na.rm=TRUE),
                    tds = sum(touchdown),
                    shortOfSticksOnThird = sum(down==3 & air_yards < ydstogo & third_down_failed, na.rm=TRUE),
                    thirdDownDropbacks = sum(down==3, na.rm=TRUE ),
                    shortOfSticksOnThirdRate = shortOfSticksOnThird / thirdDownDropbacks,  
                    passYards = sum(yards_gained[pass_attempt==1], na.rm=TRUE),
                    rushYards = sum(yards_gained[pass_attempt==0], na.rm=TRUE),
                    team = last(posteam),
                    throwAways = sum(is.na(receiver_player_name) & pass_attempt==1),
                    throwAwayRate = throwAways/numDropbacks,
                    
                    #EPA
                    meanEPA = mean(epa),
                    sdEPA = sd(epa)
                    ) %>% 
           filter(passAttempts > 200) %>%
           arrange(desc(meanEPA)) %>% mutate(ID = paste(QB, "_", year))


head(Stats)
```


Okay so now lets combine all these into one "gunslinger" score. We'll standardize each variable so it has a mean of 0 and a standard deviation of 1, using the scale() command. I then weight each feature by how important I think this is:

TD passes - weight 75%
Deep Passes (average air yards) - weight 100%
Scramble Yards - weight 25%
Not throwing short of Sticks on Third - weight 75%
Lots of pass yards - weight 75%


```{r}
Gunslinger <- Stats %>% select(QB, year, tds, passYards, meanAirYards, rushYards, shortOfSticksOnThirdRate, team)
cols <- c('tds', 'passYards', 'rushYards', 'shortOfSticksOnThirdRate', 'meanAirYards')
tdWeight <- 0.75
passWeight <- 0.5
rushWeight <- 0.25
shortWeight <- 0.75
airWeight <- 1
Gunslinger[cols] <- scale(Gunslinger[cols])
Gunslinger <- Gunslinger %>% mutate(GunslingScore = tdWeight* tds + passWeight* passYards + rushWeight* rushYards - shortWeight * shortOfSticksOnThirdRate + airWeight* meanAirYards) %>%  arrange(-GunslingScore)
Gunslinger$GunslingScore <- scale(Gunslinger$GunslingScore)

```


And now for game managers:

Don't throw interceptions - 100%
High success rate (no negative plays) - 70%
Don't take sacks - 50%
High completion rate - 75%


```{r}
Manager <- Stats %>% select(QB, year, ints, successRate, sacks, completionPercent, team, meanEPA, sdEPA)
cols <- c('ints', 'successRate', 'sacks', 'completionPercent')
Manager[cols] <- scale(Manager[cols])
intWeight <- 1
successWeight <- 0.70
sacksWeight <- 0.50
completeWeight <- 75
Manager <- Manager %>% mutate(ManageScore = -1*ints*intWeight + successRate*successWeight - sacks*sacksWeight + completionPercent*completeWeight) %>%  arrange(-ManageScore) 
Manager$ManageScore <- scale(Manager$ManageScore)

```



Now we joing them together and plot them. To keep the graph from getting cluttered, I'm just plotting 2019. 

Looking at this graph, quarterbacks in the upper left are true "gunslingers" - high performance as gunslingers, and low performance as game managers. Players in the bottom right are pure game managers. Quarterbacks in the upper right are just really good, while those in the lower left are just not very good at anything.  

Jameis Winston is a great gunslinger, but a terrible game manager, and to a lesser extent, so was Aaron Rodgers. Drew Brees, however, is a great game manager, and not much of a gunslinger anymore. In the upper right, Pat Mahomes is really good at everything.  


```{r, layout="l-page", fig.height=5.0}
library(nflfastR)
library(ggrepel)
library(ggimage)



Combined <- Gunslinger %>% left_join(Manager) %>%  left_join(teams_colors_logos, by = c('team' = 'team_abbr')) %>% arrange(-GunslingScore)

Combined %>% filter(year == 2019) %>%  ggplot(aes(x=ManageScore, y=GunslingScore, label=QB)) +
  geom_image(aes(image = team_logo_espn)) +
  geom_text_repel(aes(label=QB)) + 
  geom_hline(yintercept = 0, color = "red", linetype = "dashed", alpha=0.5) +
  geom_vline(xintercept =  0, color = "red", linetype = "dashed", alpha=0.5) + 
  labs(x = "Game Managing Performance", y="Gunslinging Performance")
  
  

```


If you disagree with my weights, I have a shiny app where you can change them and generate your own scores: https://davidranderson.shinyapps.io/GunslingersVsGameManagers/



<!-- ####################################################################### -->
<!-- Place at end of document 
     Please keep thisc hunk as is at end of your document. 
     It will create a hyperlink to the source file. -->

```{r gh-source, results='asis', echo=FALSE}
'%>%' <- magrittr::`%>%`
fld <- fs::path_wd() %>% fs::path_split() %>% purrr::pluck(1) %>% tibble::as_tibble() %>% dplyr::slice_tail(n = 1)
fn <- fs::path_wd() %>% fs::dir_ls() %>% fs::path_filter("*.Rmd") %>% fs::path_rel()
glue::glue('<a href="https://github.com/mrcaseb/open-source-football/blob/master/_posts/{fld}/{fn}"
               style="font-family:Consolas;color:blue;background-color:#f8f8f8;align:right;font-size:75%;"
              >View source code on GitHub
           </a>'
           )
```

<!-- ####################################################################### -->
