---
title: "Wins Above Expectation"
description: |
  This article looks at the percentage of snaps with win probability over an 
  arbitralily chosen critical value and compares it with the true win percentage.
author:
  - name: Sebastian Carl
    url: https://twitter.com/mrcaseb
    affiliation: nflfastR
    affiliation_url: https://twitter.com/nflfastR
date: 08-19-2020
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_depth: 3
repository_url: "https://github.com/mrcaseb/analytics-community"
categories:
  - Plot
  - nflfastR
  - Win percentage
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  message = FALSE, 
  warning = FALSE, 
  dpi = 300, 
  tidy = 'styler'
  )
```

## Preface

In the NFL, practically everyone can beat anyone. So it often happens that games 
are tight until the very end and the winner is likely to have had some luck.
Every year there are teams where you subjectively feel that they have lost or 
won particularly many of the aforementioned games.

In this post I will show off a very simple way to illustrate that by looking at
how many snaps a team played with the [nflfastR win probability](https://mrcaseb.github.io/nflfastR/articles/nflfastR-models.html) 
above a critical value (50%) more or less arbitrarily chosen by me and compare 
this value with the true win percentage.

## Load nflfastR Play by Play and compute some helper columns

Since we want to compute true win percentage from nflfastR play-by-play data
we have to do a little data wrangling before we can create the plot.

```{r, layout="l-page"}
library(tidyverse)

# Parameter --------------------------------------------------------------------

season <- 2019
wp_limit <- 0.5

# Load the data ----------------------------------------------------------------

pbp <- readRDS(url(
  glue::glue("https://raw.githubusercontent.com/guga31bb/nflfastR-data/master/data/play_by_play_{season}.rds")
  )) %>% 
  filter(pass == 1 | rush == 1)

# Compute outcomes and win percentage ------------------------------------------

outcomes <- pbp %>%
  group_by(season, game_id, home_team) %>%
  summarise(
    home_win = if_else(result > 0, 1, 0),
    home_tie = if_else(result == 0, 1, 0)
  ) %>%
  group_by(season, home_team) %>%
  summarise(
    home_games = n(),
    home_wins = sum(home_win),
    home_ties = sum(home_tie)
  ) %>%
  ungroup() %>% 
  left_join(
    # away games
    pbp %>% 
      group_by(season, game_id, away_team) %>% 
      summarise(
        away_win = if_else(result < 0, 1, 0),
        away_tie = if_else(result == 0, 1, 0)) %>%
      group_by(season, away_team) %>% 
      summarise(
        away_games = n(), 
        away_wins = sum(away_win),
        away_ties = sum(away_tie)
      ) %>% 
      ungroup(),
    by = c("season", "home_team" = "away_team")
  ) %>%
  rename(team = "home_team") %>%
  mutate(
    games = home_games + away_games,
    wins = home_wins + away_wins,
    losses = games - wins,
    ties = home_ties + away_ties,
    win_percentage = (wins + 0.5 * ties) / games
  ) %>%
  select(
    season, team, games, wins, losses, ties, win_percentage
  ) 

# Compute percentage of plays with wp > wp_lim ---------------------------------

wp_combined <- pbp %>% 
  filter(!is.na(vegas_wp) & !is.na(posteam)) %>% 
  group_by(season, posteam) %>% 
  summarise(
    pos_plays = n(),
    pos_wp_lim_plays = sum(vegas_wp > wp_limit)
  ) %>% 
  ungroup() %>% 
  left_join(
    pbp %>% 
      filter(!is.na(vegas_wp) & !is.na(posteam)) %>% 
      group_by(season, defteam) %>% 
      summarise(
        def_plays = n(),
        def_wp_lim_plays = sum(vegas_wp < wp_limit)
      ) %>% 
      ungroup(),
    by = c("season", "posteam" = "defteam")
  ) %>%
  rename(team = "posteam") %>%
  mutate(
    wp_lim_percentage = as.numeric(pos_wp_lim_plays + def_wp_lim_plays) / as.numeric(pos_plays + def_plays)
  ) %>% 
  select(season, team, wp_lim_percentage)

# Combine data and add colors and logos ----------------------------------------

chart <- outcomes %>% 
  left_join(wp_combined, by = c("season", "team")) %>% 
  filter(!is.na(wp_lim_percentage)) %>% 
  mutate(diff = 100 * (win_percentage - wp_lim_percentage)) %>% 
  group_by(team) %>% 
  summarise_all(mean) %>% 
  ungroup() %>% 
  inner_join(
    nflfastR::teams_colors_logos %>% select(team_abbr, team_color, team_logo_espn, team_logo_wikipedia),
    by=c('team' = 'team_abbr')
  ) %>% 
  mutate(
    grob =  map(seq_along(team_logo_espn), function(x) {
    grid::rasterGrob(magick::image_read(team_logo_espn[[x]]))
      })
  ) %>% 
  select(team, win_percentage, wp_lim_percentage, diff, team_color, grob) %>% 
  arrange(desc(diff))
```

## Create the plots

We will create two separate plots. A scatterplot comparing true win percentage
with the percentage of plays with win probability > 50% and a barplot showing 
the difference between the above variables.

```{r, layout="l-page", fig.height=5}
# Create scatterplot -----------------------------------------------------------
chart %>% 
  ggplot(aes(x = wp_lim_percentage, y = win_percentage)) +
  geom_abline(intercept = 0, slope = 1) +
  geom_hline(aes(yintercept = mean(win_percentage)), color = "red", linetype = "dashed") +
  geom_vline(aes(xintercept = mean(wp_lim_percentage)), color = "red", linetype = "dashed") +
  ggpmisc::geom_grob(aes(x = wp_lim_percentage, y = win_percentage, label = grob), vp.width = 0.05) +
  labs(
    x = glue::glue("Percentage of snaps with win probability over {100 * wp_limit}%"),
    y = "True win percentage (including ties as half a win)",
    title = "NFL Team Efficiency",
    caption = "Figure: @mrcaseb | Data: @nflfastR"
  ) +
  ggthemes::theme_stata(scheme = "sj", base_size = 8) +
  theme(
    plot.title = element_text(face = 'bold'),
    plot.caption = element_text(hjust = 1),
    axis.text.y = element_text(angle = 0, vjust = 0.5),
    legend.title = element_text(size = 8, hjust = 0, vjust = 0.5, face = 'bold'),
    legend.position = "top",
    aspect.ratio = 1/1.618
  ) +
  NULL
```

```{r, layout="l-page", fig.height=5, preview=TRUE}
# Create bar plot  -------------------------------------------------------------
chart %>% 
  ggplot(aes(x = seq_along(diff), y = diff)) +
  geom_hline(aes(yintercept = mean(diff)), color = "red", linetype = "dashed") +
  geom_col(width = 0.5, colour = chart$team_color, fill = chart$team_color, alpha = 0.5) +
  ggpmisc::geom_grob(aes(x = seq_along(diff), y = diff, label = grob), vp.width = 0.035) +
  # scale_x_continuous(expand = c(0,0)) +
  labs(
    x = "Rank",
    y = "Win Percentage Over Expectation",
    title = "NFL Team Efficiency",
    subtitle = "How Lucky are the Teams?",
    caption = "Figure: @mrcaseb | Data: @nflfastR"
  ) +
  ggthemes::theme_stata(scheme = "sj", base_size = 8) +
  theme(
    plot.title = element_text(face = 'bold'),
    plot.caption = element_text(hjust = 1),
    axis.text.y = element_text(angle = 0, vjust = 0.5),
    legend.title = element_text(size = 8, hjust = 0, vjust = 0.5, face = 'bold'),
    legend.position = "top",
    aspect.ratio = 1/1.618
  ) +
  NULL
```
