---
title: "Evauluating defenses by how well they play the offense's WR 1"
description: |
  Cornerback performance is hard to measure, but we can try.
author:
  - name: Jonas Trostle
    url: https://twitter.com/JonasTrostle/
date: "`r Sys.Date()`"
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_depth: 3
repository_url: "https://github.com/mrcaseb/open-source-football"
categories:
  - Figures
  - nflfastR
  - Defense
---

<!-- ####################################################################### -->
<!-- Please keep the following chunk as is at the top of your document. 
It will set some global chunk options.  -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  message = FALSE, 
  warning = FALSE, 
  dpi = 300, 
  tidy = 'styler'
)
```

<!-- ####################################################################### -->

Roll the clocks back to 2009. Tom Brady is back from tearing his ACL and MCL, Brett Favre is a Viking, and the Colts and Saints are a combined 26-0 by week 13. Nothing was quite as beautiful as what Darrelle Revis was doing for the New York Jets. Week after week, Revis would eliminate the opponent's best receiver. 

Measuring this dominance is hard however. Can we do better than just adding up the EPA from his swats and tackles, or just summarizing the defense as a whole? Instead of that, what if we ranked every team's wide receivers and compared how each defense did against the opponent's number one playmaker?

## Data & transformations

First we need to load in the data, and to do that we follow Thomas Mock's great guide on using Arrow to speed up the process. 

```{r}
library(arrow)
library(tidyverse)
library(ggrepel)
library(nflfastR)
library(ggpmisc)
options(mc.cores = parallel::detectCores())

set.seed(2009)

dir.create("nflfastr")

get_data <- function(year){
  dir.create(file.path("nflfastr", year))

  download.file(
    glue::glue("https://github.com/guga31bb/nflfastR-data/blob/master/data/play_by_play_{year}.parquet?raw=true"),
    file.path("nflfastr", year, "data.parquet"),
    mode = 'wb'
  )
}

walk(1999:2019, get_data)

ds <- open_dataset("nflfastr", partitioning = "year")
# We're grabbing everything since 1999 so I only have to do it once,
# but feel free to just do individual seasons
ds %>%
  select(
    desc,
    posteam,
    defteam,
    receiver,
    receiver_id,
    epa,
    cpoe,
    pass,
    qb_epa,
    air_yards,
    year,
    yards_gained,
    penalty
  ) %>%
  filter(year >= 1999, penalty == 0,!is.na(epa)) %>%
  collect() -> pbp
```

Next, we want to combine the play by play data with the roster data for receivers.

```{r}
# Again, if you're just doing an individual season you can change this
nflfastR::fast_scraper_roster(1999:2020) -> roster

pbp %>%
  select(
    desc,
    posteam,
    defteam,
    receiver,
    receiver_id,
    epa,
    cpoe,
    pass,
    qb_epa,
    air_yards,
    year,
    yards_gained
  ) %>%
  filter(!is.na(receiver_id)) -> pbp2

pbp2 %>%
  left_join(roster, by = c("receiver_id" = "gsis_id")) -> pbp3
```

Now that we have the play by play merged with the roster data, we can go ahead and start transforming the data. We need to rank an offense's wide receivers, and there are many ways to do so. For now, we'll use targeted air yards, but we can change this decision later.

```{r}
pbp3 %>%
  # we keep only the wide receivers, no tight ends or running backs
  filter(position == "WR") %>% 
  group_by(receiver_id, year) %>%
  # we sum up all the receivers air yards when targeted
  mutate(targeted_air_yards = sum(air_yards)) %>%
  ungroup() %>%
  # distinct allows us to keep only one row per player
  distinct(receiver_id, year, .keep_all = T) %>%
  # we want to rank receivers by team, not overall, so we group by offense
  group_by(posteam, year) %>%
  # since we want to rank the receivers, arrange allows us to order them by some
  # criterion
  arrange(-targeted_air_yards) %>%
  # this index is now the within-team rank of each wide receiver that year
  mutate(index = row_number()) %>%
  ungroup() %>%
  select(receiver_id, targeted_air_yards, index, year) -> pbp4
```

Now that we have every receivers rank, we'll add that back to the play by play data and remove any plays that didn't have a receiver targeted.

```{r}
pbp2 %>%
  left_join(pbp4) -> pbp5

pbp5 %>%
  filter(!is.na(index)) -> pbp6
```

At this point, I think the best way to graph this would be to compare how a defense does against WR 1, and against all other wide receivers. To do this, we'll create a binary variable that says whether or not a receiver was number one. Uncreatively, I called this the smittywerbenjagermenjensen index, but you can rename it whatever you like.

```{r}
pbp6 %>%
  mutate(smittywerbenjagermenjensen = if_else(index == 1, 1,0)) -> pbp7
```

Almost done. We're going to group by the new smittywerbenjagermenjensen index and by defense and calculate the average EPA given up to WR 1 vs all others. Like ranking the receivers, there are lots of ways to skin a cat: possible options include using qb_epa instead, or CPOE, or summing any of the above instead of averaging them. All will give a slightly different answer.

```{r}
pbp7 %>%
  group_by(defteam, smittywerbenjagermenjensen, year) %>%
  summarise(epa = mean(epa)) %>%
  ungroup() -> pbp8

# I could live to be 100 and never learn how to pivot correctly on the first try

pbp8 %>%
  pivot_wider(names_from = smittywerbenjagermenjensen,
              values_from =  epa,
              names_prefix = "wr_") -> pbp9

# We'll add the team colors so the graph looks nice

pbp9 %>%
  left_join(teams_colors_logos, by = c('defteam' = 'team_abbr')) -> pbp10

pbp10 %>%
  filter(year == 2009) -> pbp11
```

## The plot (thickens)


Is it done? Yes! Now we can plot it and see which defenses could shutdown each opponent's best receiver.

```{r "", echo = FALSE, layout="l-body-outset", fig.height=6}
pbp11 %>%
  ggplot(aes(x = wr_0, y = wr_1)) +
  geom_point(color = pbp11$team_color) +
  geom_hline(
    yintercept = mean(pbp11$wr_1),
    color = "red",
    linetype = "dashed",
    alpha = 0.5
  ) +
  geom_vline(
    xintercept =  mean(pbp11$wr_0),
    color = "red",
    linetype = "dashed",
    alpha = 0.5
  ) +
  geom_text_repel(aes(label = defteam)) +
  labs(
    x = "EPA allowed to all other wide receivers",
    y = "EPA allowed to WR 1",
    title = "Defensiveback Efficiency 2009",
    caption = "Data: @nflfastR"
  ) +
  theme_bw() +
  theme(plot.title = element_text(size = 14, hjust = 0.5, face = "bold")) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 10)) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) +
  scale_y_reverse() +
  scale_x_reverse()
```
Wow! At a glance we can see that there are three teams that really shutdown the opponent's best wide receiver. If you think back to '09, even though Revis founded his island that year and solidified himself as the best cover corner in the league, it was Charles Woodson of the Packers who won defensive player of the year. In a very strange twist of fate however, the best performance was from New Orleans. 

The coolest part about this, for me at least, is that we could recover all of these narratives from the data. Without measuring any cornerbacks directly, we can get a measurement that passes the eye test rather convincingly.

Of course, this measure is flawed in that we still can't directly measure a player's contribution. For example, the corners in New Orleans almost always had Darren Sharper over top to help him, and there's no guarantee that they were actually covering WR 1 every play. For a player like Revis, you could argue that because he could eliminate the opponent's WR 1 the rest of the defense could roll towards the other receivers and play them better, which is why the Jets are so much more rightward on the graph than the Saints.

## Coda

We could argue endlessly about the best cover corners and seasons in NFL history, but as a coherence check I think we should take a look at how this new metric ranks every defense since 1999. To paraphrase Ben Baldwin, a coherence check for these models is "Do we find that Stephon Gilmore is in the top 5?"

With that in mind, let's check out what the model thinks the best coverage season is. The code is mostly identical to the first section, but I put comments throughout the code to explain any changes. The biggest change is that we can't use targeted air yards, since those don't go far enough back, so I've elected to use total yards. Again, total EPA would be just as legitimate, as would targets or receptions. I invite you to play around with it.

```{r "", echo = FALSE, layout="l-body-outset", fig.height=6}
pbp3 %>%
  filter(position == "WR") %>%
  group_by(receiver_id, year) %>%
  # since we don't have air yards back to 1999, this time we'll use raw yards
  mutate(total_yards_gained = sum(yards_gained)) %>%
  ungroup() %>%
  distinct(receiver_id, year, .keep_all = T) %>%
  group_by(posteam, year) %>%
  # arrange by the new criterion
  arrange(-total_yards_gained) %>%
  mutate(index = row_number()) %>%
  ungroup() %>%
  # and remember to grab the new one
  select(receiver_id, total_yards_gained, index, year) -> pbp4

pbp2 %>%
  left_join(pbp4) -> pbp5

pbp5 %>%
  filter(!is.na(index)) -> pbp6

pbp6 %>%
  mutate(smittywerbenjagermenjensen = if_else(index == 1, 1, 0)) -> pbp7

pbp7 %>%
  group_by(defteam, smittywerbenjagermenjensen, year) %>%
  summarise(epa = mean(epa)) %>%
  ungroup() -> pbp8

pbp8 %>%
  pivot_wider(names_from = smittywerbenjagermenjensen,
              values_from =  epa,
              names_prefix = "wr_") -> pbp9

pbp9 %>%
  left_join(teams_colors_logos, by = c('defteam' = 'team_abbr')) -> pbp10

# we need name and year labels instead of just team name, so we create that here
pbp10 %>%
  mutate(team_label = str_c(defteam, year, sep = " ")) -> pbp10
```

```{r "", echo = FALSE, layout="l-body-outset", fig.height=6}
pbp10 %>%
  ggplot(aes(x = wr_0, y = wr_1)) +
  geom_point(color = pbp10$team_color) +
  geom_hline(
    yintercept = mean(pbp10$wr_1),
    color = "red",
    linetype = "dashed",
    alpha = 0.5
  ) +
  geom_vline(
    xintercept =  mean(pbp10$wr_0),
    color = "red",
    linetype = "dashed",
    alpha = 0.5
  ) +
  # We label only the least dense 1.5%, which should give us the outliers
  stat_dens2d_filter(geom = "text_repel",
                     keep.fraction = 0.015,
                     # we use the new team labels we created
                     aes(label = team_label)) +
  labs(
    x = "EPA allowed to all other wide receivers",
    y = "EPA allowed to WR 1",
    title = "Defensiveback Efficiency, 1999-2020",
    caption = "Data: @nflfastR"
  ) +
  theme_bw() +
  theme(plot.title = element_text(size = 14, hjust = 0.5, face = "bold")) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 10)) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) +
  scale_y_reverse() +
  scale_x_reverse()
```

Sure enough, Stephon Gilmore's 2019 season is in the top right, and ranks #4 since 1999. Surprisingly, that 2009 season by the Saints ranks  as the best of the millennium. Just as surprising is that the Legion of Boom is nowhere in sight. 

For the astute viewer, you may notice that the coverage score against WR 1 and against all other WRs is correlated. This could lend credibility to the theory that a shutdown corner allows the defense to divert resources elsewhere, or it could just be that good coordinators get better performances from all corners.

This is by no means the be-all end-all of cornerback rankings. I'd love to see this extended to account for the quality of receivers faced and for different eras. A version looking at all receivers, including TEs and RBs, may be even more enlightening. Finally, like the team tiers, average EPA allowed to WR 1 is worth a different amount than the average EPA allowed to all other receivers, and it would be great to sort these defenses into tiers, as well as do a stability analysis to see if any of these measures are more stable than regular defense.


<!-- ####################################################################### -->
<!-- Place at end of document 
Please keep this chunk as is at end of your document. 
It will create a hyperlink to the source file. -->

```{r gh-source, results='asis', echo=FALSE}
'%>%' <- magrittr::`%>%`
fld <- fs::path_wd() %>% fs::path_split() %>% purrr::pluck(1) %>% tibble::as_tibble() %>% dplyr::slice_tail(n = 1)
fn <- fs::path_wd() %>% fs::dir_ls() %>% fs::path_filter("*.Rmd") %>% fs::path_rel()
glue::glue('<a href="https://github.com/mrcaseb/open-source-football/blob/master/_posts/{fld}/{fn}"
               style="font-family:Consolas;color:blue;background-color:#f8f8f8;align:right;font-size:75%;"
              >View source code on GitHub
           </a>'
)
```

<!-- ####################################################################### -->
