---
title: "Python contributing example"
description: |
  Showing how to contribute using Python code
author:
  - name: Ben Baldwin
    url: https://twitter.com/benbbaldwin
date: 08-20-2020
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_depth: 3
repository_url: "https://github.com/mrcaseb/open-source-football"
categories:
  - nflfastR
  - python
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  message = FALSE, 
  warning = FALSE, 
  dpi = 300, 
  tidy = 'styler'
  )

library(reticulate)
```

## Contributing in python

### There is some stuff to install and setup:

* R and RStudio
* Get github in RStudio setup ([see here](http://r-pkgs.had.co.nz/git.html#git-setup))
* In R, run `install.packages(c("distill", "rmarkdown"))` to get the needed packages for the website installed, and `install.packages("reticulate")` to install `reticulate`, which allows python to be run from within R
* You'll need to somehow point RStudio to your python installation. I ran `py_install("pandas")` and the below worked

### Once setup is done:

Start working on your document. You can put in python code chunks as in the examples below ([see here](https://github.com/mrcaseb/open-source-football/blob/master/_posts/2020-08-20-python-contributing-example/python-contributing-example.Rmd) for the source code).

* Once you have everything ready in your document, press the "Knit" button on the top and it will create the document.
* Then you can committ all the files using the Git tab
* Please see the [step-by-step instructions here](https://mrcaseb.github.io/open-source-football/contributing.html#more-detailed-instruction)

Once everything is set up, you can just use python code as normal. Here is a quick example borrowing some code from [from Deryck's nflfastR python guide](https://gist.github.com/Deryck97/dff8d33e9f841568201a2a0d5519ac5e).

```{python p1}
import pandas
data = pandas.read_csv("https://github.com/guga31bb/nflfastR-data/blob/master/data/play_by_play_2019.csv.gz?raw=True", compression='gzip', low_memory=False)
data = data.loc[(data.play_type.isin(['no_play','pass','run'])) & (data.epa.isna()==False)]
data.groupby('posteam')[['epa']].mean()
```

Grouping by QBs:

```{python p2}
qbs = data.groupby(['passer','posteam'], as_index=False).agg({'epa':'mean',
                                                              'cpoe':'mean',
                                                              'play_id':'count'})
# at least 200 plays
qbs = qbs.loc[qbs.play_id>199]
# sort by EPA
qbs.sort_values('epa', ascending=False, inplace=True)

#Round to two decimal places where appropriate
qbs = qbs.round(2)

#Rename columns
qbs.columns = ['Player','Team','EPA per Dropback','CPOE','Dropbacks']

qbs
```

Hopefully the process is painless once all the setup is done.

```{r gh-source, results='asis', echo=FALSE}
'%>%' <- magrittr::`%>%`
fld <- fs::path_wd() %>% fs::path_split() %>% purrr::pluck(1) %>% tibble::as_tibble() %>% dplyr::slice_tail(n = 1)
fn <- fs::path_wd() %>% fs::dir_ls() %>% fs::path_filter("*.Rmd") %>% fs::path_rel()
glue::glue('<a href="https://github.com/mrcaseb/open-source-football/blob/master/_posts/{fld}/{fn}"
               style="font-family:Consolas;color:blue;background-color:#f8f8f8;align:right;font-size:75%;"
              >View source code on GitHub
           </a>'
           )
```
