---
title: "Visualizing EPSN's Total QBR Using Interactive Plots"
description: |
  A short description of the post.
author:
  - name: Sebastian Carl
    url: https://twitter.com/mrcaseb
    affiliation: nflfastR
    affiliation_url: https://twitter.com/nflfastR
date: 08-24-2020
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_depth: 3
repository_url: "https://github.com/mrcaseb/open-source-football"
categories:
  - Scraping
  - espnscrapeR
  - Total QBR
  - Figures
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  message = FALSE, 
  warning = FALSE, 
  dpi = 300, 
  tidy = 'styler',
  layout = "l-page"
  )
```

## Preface

When I first started working with ESPN's Total QBR data I developed my own code 
to scrape the website. For this post I dug out the old (bad) code and wanted to 
make it prettier. But at the last minute I remembered something: there is this
undervalued R package [`espnscrapeR`](https://github.com/jthomasmock/espnscrapeR)
developed and maintained by [Thomas Mock](https://twitter.com/thomas_mock).
So instead of implementing the scraper by myself in this post I'll show off,
how to get all the data and save it locally for later reuse.

And as the title suggests we want to try out something new here: interactive plots

## How to get the data

We want to use the `espnscrapeR` in this section. So let's install it by running
the following code block.

```{r, eval=FALSE}
if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}
remotes::install_github("jthomasmock/espnscrapeR")
```

Now that `espnscrapeR` is installed we will load weekly Total QBR. 

We'll do this by writing the function `get_week_qbr` which loads the QBR data 
for a given week and season.

```{r, eval=FALSE}
get_week_qbr <- function(s, wk) {
  if (wk <= 17) {
    out <- espnscrapeR::get_nfl_qbr(season = s, season_type = "Regular", week = wk)
  } else if (wk > 17 & s != 2017) {
    out <- espnscrapeR::get_nfl_qbr(season = s, season_type = "Playoffs", week = wk - 17)
  }
}
```

Then we call this function to actually do the job and save the output. Note: We 
call the function for the weeks 1 to 21 although the SB is week 22 because of the
Pro Bowl. The package automatically converts to week 22 for us.

The code is somewhat advanced and You may need to do some more background 
reading to understand it. The short version: we call the above function for all
seasons between 2006 and 2019 and all available weeks in parallel running processes 
and bind everything together in a dataframe which will saved to disc for further
usage.

```{r, eval=FALSE}
library(dplyr)
future::plan("multiprocess")

# Total QBR data as available beginning in 2006
# Note: Playoffs 2017 are missing 
seasons <- 2006:2019 

all_qbr <- 
  furrr::future_pmap_dfr(purrr::transpose(purrr::cross2(seasons, 1:21)), get_week_qbr) %>% 
  dplyr::mutate(
    game_week = as.numeric(game_week),
    game_week = dplyr::if_else(season_type == "Playoffs", game_week + 17, game_week)
  ) %>% 
  dplyr::arrange(season, game_week)

# save to disk------------------------------------------------------------------
# binary
saveRDS(all_qbr, file = "all_qbr.rds")

# ASCII
readr::write_csv(all_qbr, "all_qbr.csv")
```

## Create some static weekly QBR plots

Let's have a look at the data by plotting weekly QBR for some selected QBs (the 
insane numbers). We want to plot multiple QBs so let's create a function again.

Before we do this please note there are some Quarterbacks appearing with special
team names (e.g. `"DEN/KC"`) in the data because they changed their team within 
a season. Our new function won't work for those QBS but I was too lazy to rewrite
it because I honestly don't care about those QBs. So please be careful when trying
to run the code with one of the following QB names and corresponding seasons.

```{r, layout="l-body"}
library(tidyverse)
readRDS("all_qbr.rds") %>% 
  dplyr::filter(stringr::str_length(team) > 3) %>% 
  dplyr::group_by(season, short_name, team) %>% 
  dplyr::summarise() %>% 
  knitr::kable("simple", align = "c")
```

Now let's write the plotting function

```{r}
plot_weekly_qbr <- function(year, QB, all_qbr_file) {
  # Filter the QB we want to look at and modify team names and week numbers
  # to fit nflfastR convention
  single_qbr <- all_qbr_file %>%
    dplyr::filter(season == year & name == QB) %>%
    dplyr::mutate(
      team = dplyr::case_when(
        team == "LAR" ~ "LA",
        team == "WSH" ~ "WAS",
        TRUE ~ team
      ),
      game_week = dplyr::if_else(game_week > 21, 21, game_week)
    ) %>%
    dplyr::left_join(
      nflfastR::teams_colors_logos %>% select(team_abbr, team_color),
      by = c("team" = "team_abbr")
    )

  # Search for the QBs opponents
  opponents <- nflfastR::fast_scraper_schedules(year) %>%
    dplyr::filter(home_team == single_qbr$team | away_team == single_qbr$team) %>%
    dplyr::mutate(
      opp = dplyr::if_else(home_team == single_qbr$team, away_team, home_team)
    ) %>%
    dplyr::left_join(
      nflfastR::teams_colors_logos %>% select(team_abbr, team_logo_espn),
      by = c("opp" = "team_abbr")
    ) %>%
    dplyr::mutate(
      grob = purrr::map(seq_along(team_logo_espn), function(x) {
        grid::rasterGrob(magick::image_read(team_logo_espn[[x]]))
      })
    )

  chart <- single_qbr %>%
    dplyr::left_join(opponents, by = c("game_week" = "week"))
  
  if (max(chart$game_week) > 17){
    title_string <- glue::glue("{QB} Weekly Total QBR {year} including Playoffs")
  } else {
    title_string <- glue::glue("{QB} Weekly Total QBR {year} Regular Season")
  }
  
  # going to draw some quantile lines and combine them here
  quantiles <- c(
    quantile(all_qbr_file$qbr_total, 0.10),
    quantile(all_qbr_file$qbr_total, 0.25),
    quantile(all_qbr_file$qbr_total, 0.75),
    quantile(all_qbr_file$qbr_total, 0.90),
    quantile(all_qbr_file$qbr_total, 0.98)
  )

  chart %>%
    ggplot(aes(x = game_week, y = qbr_total)) +
    geom_hline(yintercept = quantiles, color = "black", linetype = "dashed", alpha = 0.7) +
    geom_hline(yintercept = quantile(all_qbr_file$qbr_total, 0.50), color = "black", linetype = "solid", alpha = 0.7) +
    geom_text(x = 0, y = 2 + quantile(all_qbr_file$qbr_total, 0.10), label = "10th Percentile", hjust = 1, size = 2) +
    geom_text(x = 0, y = 2 + quantile(all_qbr_file$qbr_total, 0.25), label = "25th Percentile", hjust = 1, size = 2) +
    geom_text(x = 0, y = 2 + quantile(all_qbr_file$qbr_total, 0.50), label = "50th Percentile", hjust = 1, size = 2) +
    geom_text(x = 0, y = 2 + quantile(all_qbr_file$qbr_total, 0.75), label = "75th Percentile", hjust = 1, size = 2) +
    geom_text(x = 0, y = 2 + quantile(all_qbr_file$qbr_total, 0.90), label = "90th Percentile", hjust = 1, size = 2) +
    geom_text(x = 0, y = 2 + quantile(all_qbr_file$qbr_total, 0.98), label = "98th Percentile", hjust = 1, size = 2) +
    geom_line(colour = chart$team_color) +
    xlim(-1.4, NA) +
    ggpmisc::geom_grob(aes(x = game_week, y = qbr_total, label = grob), vp.width = 0.05) +
    labs(
      x = "Game Week",
      y = "ESPN Total QBR",
      caption = "Figure: @mrcaseb | Data: espnscrapeR by @thomas_mock",
      title = title_string,
      subtitle = "Percentiles of the whole NFL 2006 - 2019 are drawn for orientation\nTo qualify, a player must play a minimum of 20 action plays per team game"
    ) +
    ggthemes::theme_stata(scheme = "sj", base_size = 8) +
    theme(
      plot.title = element_text(face = "bold"),
      plot.caption = element_text(hjust = 1),
      axis.text.y = element_text(angle = 0, vjust = 0.5),
      legend.title = element_text(size = 8, hjust = 0, vjust = 0.5, face = "bold"),
      legend.position = "top",
      aspect.ratio = 1 / 1.618
    ) +
    NULL
}
```

Well, that's a lot of code. Let's see if it is working properly by choosing the 
insane 2007 season of Tom Brady (in terms of Total QBR the all-time best season).

```{r, fig.height=5, preview=TRUE}
all_qbr <- readRDS("all_qbr.rds")
plot_weekly_qbr(2007, "Tom Brady", all_qbr)
```

Ok let's look at the other top 3 seasons

```{r, fig.height=5}
plot_weekly_qbr(2006, "Peyton Manning", all_qbr)
```

```{r, fig.height=5}
plot_weekly_qbr(2011, "Aaron Rodgers", all_qbr)
```

And what does it look like if the season wasn't that good?

```{r, fig.height=5}
plot_weekly_qbr(2019, "Mitchell Trubisky", all_qbr)
```


## Let's make it interactive

Those plots are nice and all but actually we may miss a lot of information
when looking at them. For example we might want to know if the game was home or
away, what the result was or how many QB plays were counted for Total QBR.

```{r}
interactive_weekly_qbr <- function(year, QB, all_qbr_file) {
  # Filter the QB we want to look at and modify team names and week numbers
  # to fit nflfastR convention
  single_qbr <- all_qbr_file %>%
    dplyr::filter(season == year & name == QB) %>%
    dplyr::mutate(
      team = dplyr::case_when(
        team == "LAR" ~ "LA",
        team == "WSH" ~ "WAS",
        TRUE ~ team
      ),
      game_week = dplyr::if_else(game_week > 21, 21, game_week)
    ) %>%
    dplyr::left_join(
      nflfastR::teams_colors_logos %>% select(team_abbr, team_color),
      by = c("team" = "team_abbr")
    )

  # Search for the QBs opponents
  opponents <- nflfastR::fast_scraper_schedules(year) %>%
    dplyr::filter(home_team == single_qbr$team | away_team == single_qbr$team) %>%
    dplyr::mutate(
      opp = dplyr::if_else(home_team == single_qbr$team, away_team, home_team),
      loc_desc = dplyr::if_else(
        home_team == single_qbr$team,
        glue::glue("vs. {away_team}"),
        glue::glue("@ {home_team}")
      ),
      game_desc = dplyr::case_when(
        week == 18 ~ "Wild Card",
        week == 19 ~ "Divisional Round",
        week == 20 ~ "Conference Championship",
        week == 21 ~ "Super Bowl",
        TRUE ~ "Regular Season"
      )
    ) %>%
    dplyr::left_join(
      nflfastR::teams_colors_logos %>% select(team_abbr, opp_color = team_color, team_logo_espn),
      by = c("opp" = "team_abbr")
    )

  chart <- single_qbr %>%
    dplyr::left_join(opponents, by = c("game_week" = "week")) %>%
    dplyr::mutate(
      outcome = dplyr::case_when(
        team == home_team & home_result > 0 ~ "Won",
        team == home_team & home_result < 0 ~ "Lost",
        team == away_team & home_result > 0 ~ "Lost",
        team == away_team & home_result < 0 ~ "Won",
        home_result == 0 ~ "Tie",
        TRUE ~ NA_character_
      )
    )


  if (max(chart$game_week) > 17) {
    title_string <- glue::glue("{QB} Weekly Total QBR {year} including Playoffs")
  } else {
    title_string <- glue::glue("{QB} Weekly Total QBR {year} Regular Season")
  }

  # going to draw some quantile lines and combine them here
  quantiles <- c(
    quantile(all_qbr_file$qbr_total, 0.10),
    quantile(all_qbr_file$qbr_total, 0.25),
    quantile(all_qbr_file$qbr_total, 0.75),
    quantile(all_qbr_file$qbr_total, 0.90),
    quantile(all_qbr_file$qbr_total, 0.98)
  )

  plot <-
    chart %>%
    ggplot(aes(x = game_week, y = qbr_total)) +
    geom_hline(yintercept = quantiles, color = "black", linetype = "dashed", alpha = 0.7) +
    geom_hline(yintercept = quantile(all_qbr_file$qbr_total, 0.50), color = "black", linetype = "solid", alpha = 0.7) +
    geom_text(x = -0.5, y = 2 + quantile(all_qbr_file$qbr_total, 0.10), label = "10th Percentile", size = 3) +
    geom_text(x = -0.5, y = 2 + quantile(all_qbr_file$qbr_total, 0.25), label = "25th Percentile", size = 3) +
    geom_text(x = -0.5, y = 2 + quantile(all_qbr_file$qbr_total, 0.50), label = "50th Percentile", size = 3) +
    geom_text(x = -0.5, y = 2 + quantile(all_qbr_file$qbr_total, 0.75), label = "75th Percentile", size = 3) +
    geom_text(x = -0.5, y = 2 + quantile(all_qbr_file$qbr_total, 0.90), label = "90th Percentile", size = 3) +
    geom_text(x = -0.5, y = 2 + quantile(all_qbr_file$qbr_total, 0.98), label = "98th Percentile", size = 3) +
    geom_line(color = chart$team_color) +
    geom_point(
      aes(text = glue::glue(
        "Week: {game_week} ({game_desc})\n{outcome} {away_score}:{home_score} {loc_desc}\n Total QBR: {qbr_total}\nNumber of QB plays: {qb_plays}"
      )),
      color = chart$opp_color,
      size = 0.05
    ) +
    xlim(-1.4, NA) +
    # ggpmisc::geom_grob(aes(x = game_week, y = qbr_total, label = grob), vp.width = 0.05) +
    labs(
      x = "Game Week",
      y = "ESPN Total QBR",
      caption = "Figure: @mrcaseb | Data: espnscrapeR by @thomas_mock",
      title = title_string,
      subtitle = "Percentiles of the whole NFL 2006 - 2019 are drawn for orientation\nTo qualify, a player must play a minimum of 20 action plays per team game"
    ) +
    theme_bw() +
    # theme(
    #   plot.title = element_text(face = "bold"),
    #   plot.caption = element_text(hjust = 1),
    #   axis.text.y = element_text(angle = 0, vjust = 0.5),
    #   legend.title = element_text(size = 8, hjust = 0, vjust = 0.5, face = "bold"),
    #   legend.position = "top",
    #   # aspect.ratio = 1 / 1.618
    # ) +
    NULL

  image_list <- chart %>%
    dplyr::transmute(
      source = team_logo_espn,
      xref = "x",
      yref = "y",
      x = game_week,
      y = qbr_total,
      sizex = 5,
      sizey = 5,
      opacity = 0.9,
      xanchor = "center",
      yanchor = "middle"
    ) %>%
    purrr::transpose()

  ggplotly(plot, tooltip = "text") %>% 
    plotly::layout(images = image_list)
}
```

```{r, fig.height=2.5}
library(plotly)
interactive_weekly_qbr(2007, "Tom Brady", all_qbr)
```

```{r, fig.height=2.5}
interactive_weekly_qbr(2019, "Mitchell Trubisky", all_qbr)
```
